---
bibliography: thesis-refs.bib
---

# Introduction

```{r}
#| label: setup
#| include: false
library(ggplot2)
library(dplyr)
library(readr)
library(scales)

fraud_fi <- read_csv(
  "data/external/statfin_fraud_offences_13ex.csv",
  show_col_types = FALSE,
  col_names = c("year", "area", "offence", "n"), skip = 1
) |>
  mutate(group = if_else(
    grepl("^111[89]", offence),
    "Means-of-payment fraud (ch. 37)",
    "Fraud (ch. 36)"
  )) |>
  summarise(n = sum(n), .by = c(year, group)) |>
  filter(year >= 1991)

fraud_since_2010 <- fraud_fi |> filter(group=="Means-of-payment fraud (ch. 37)", year %in% c(2010, max(year)))
fraud_prc_rise <- (max(fraud_since_2010$n)-min(fraud_since_2010$n))/min(fraud_since_2010$n)*100
fraud_prc_rise <- round(fraud_prc_rise,1)
```

## Fraud detection {#sec-intro-fraud}

Financial fraud is a substantial and growing problem. Police-recorded means-of-payment fraud offences in Finland have risen by `r fraud_prc_rise` % between 2010 and `r max(fraud_since_2010$year)` (@fig-fi-fraud-trend). The monetary losses follow a similar trajectory. Finance Finland, the trade association of the Finnish financial sector, reports that criminals attempted to defraud Finnish bank customers of EUR 148 million in 2025, up `r (((148-76.9)/76.9) |> round(digits = 3))*100` % from 2023; banks blocked or recovered roughly half of the attempted amount, leaving EUR 72.5 million in realised losses to customers [@finanssiala2025_huijaukset_2024; @finanssiala2026_huijaukset_2025]. Nor is the trend specific to Finland: a joint report by the European Banking Authority (EBA) and the European Central Bank (ECB) puts the total value of fraudulent payment transactions in the European Economic Area at EUR 4.2 billion in 2024, a 17 % year-on-year increase [@ebaecb2025_payment_fraud_report].

```{r}
#| label: fig-fi-fraud-trend
#| fig-cap: "Police-recorded fraud offences in Finland, 1991-2025. Fraud covers fraud, petty fraud and aggravated fraud (Criminal Code ch. 36). Means-of-payment fraud covers its petty and aggravated forms (ch. 37), criminalised as a separate offence category in 1991. [@statfin2026_fraud_offences_13ex]"
#| fig-width: 6.2
#| fig-height: 3.2
#| warning: false
#| message: false

ggplot(fraud_fi, aes(year, n, colour = group)) +
  geom_line(linewidth = 1.2) +
  scale_colour_manual(values = c(
    "Fraud (ch. 36)" = "#2563EB",
    "Means-of-payment fraud (ch. 37)" = "#D97706"
  )) +
  scale_x_continuous(breaks = seq(1980, 2025, 5)) +
  scale_y_continuous(labels = label_number(big.mark = ","), limits = c(0, NA)) +
  labs(x = NULL, y = "Offences known to the authorities", colour = NULL) +
  theme_minimal(base_size = 10) +
  theme(
    legend.position = "bottom",
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank()
  )
```

Financial fraud is commonly divided into two distinct types: first-party fraud, in which the customer themselves is the fraudster, and third-party fraud, in which the fraud is committed by an outside actor who is not a legitimate participant in the transaction. The fraud taxonomy developed by the Euro Banking Association encodes this distinction [@moes2024_pan_european_fraud_taxonomy]. This thesis concentrates on third-party fraud, and the term fraud will hereafter refer to this type.

Third-party fraud can be further divided by who initiates the fraudulent payment. European fraud reporting under PSD2 draws this line between unauthorised payment transactions, in which the fraudster issues the payment order without the payer's consent, as in account takeovers or payments made with stolen card details, and manipulation of the payer, in which the customer is misled into authorising the payment in good faith [@moes2024_pan_european_fraud_taxonomy; @ebaecb2025_payment_fraud_report]. The latter case makes detection harder still, since at the moment of the transaction even the paying customer may not know that they are being defrauded. It is also the growing case: manipulation of the payer accounted for 74 % of the value of fraudulent credit transfers in the EEA in 2024, up from 65 % in 2023, and strong customer authentication does not stop a payment the customer authorises themselves and thus there is need for payment level detection algorithms[@ebaecb2025_payment_fraud_report].

Algorithmic detection of fraudulent transactions faces two primary difficulties. The first is class imbalance: fraudulent transactions are only a small fraction of all transactions, and such severe imbalance degrades machine-learning-based models regardless of family. The second is the changing modus operandi of fraud. Detection is a two-actor system in which fraudsters abandon attack patterns as the detection systems learn them. This second difficulty is out of scope for this thesis, which concentrates on the first.

```{r}
#| label: fig-dataset-prevalence
#| fig-cap: "Fraud prevalence of the datasets used in this thesis. The dashed line marks the 0.033 % share of fraud rates reported for payment transactions in the European Economic Area in 2024 [@ebaecb2025_payment_fraud_report]. Prevalences are computed over each full dataset."
#| fig-width: 6.2
#| fig-height: 3.0
#| warning: false
#| message: false

# Prevalences extracted from data/duckdb/fraud_datasets.duckdb
# (full datasets, incl. FiFAR's pre-defined test split).
prevalence <- read_csv(
  "data/external/dataset_fraud_prevalence.csv",
  show_col_types = FALSE
) |>
  mutate(
    fraud_rate = n_fraud / n_rows,
    dataset_label = reorder(dataset_label, fraud_rate)
  )

eba_range <- c(0.00001, 0.00033)

ggplot(prevalence, aes(dataset_label, fraud_rate)) +
  geom_col(fill = "#D97706") +
  geom_text(
    aes(label = percent(fraud_rate, accuracy = 0.01)),
    vjust = -1, size = 3, colour = "grey20"
  ) +
  geom_hline(
    yintercept = eba_range[2],
    linetype = "dashed",
    size = 1
  ) +
  scale_y_continuous(
    breaks = c(0, 0.001, 0.01),
    labels = label_percent(),
    limits = c(0, 0.02)
  ) +
  labs(x = NULL, y = "Share of fraudulent transactions") +
  theme_minimal(base_size = 10) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank()
  )
```

The imbalance in openly available research datasets, severe as it is, still understates the production setting. The EBA and ECB report fraud rates by value between 0.001 % and 0.033 % across payment instrument categories, with card transactions at the top of that range [@ebaecb2025_payment_fraud_report]. The fraud prevalence of the datasets used in this thesis ranges from `r percent(min(prevalence$fraud_rate), accuracy = 0.01)` to `r percent(max(prevalence$fraud_rate), accuracy = 0.1)` (@fig-dataset-prevalence). This is one to three orders of magnitude higher. In other words, the public benchmark datasets are already heavily oversampled relative to the class distribution a deployed detection system faces.

Under such imbalance, classification accuracy is a meaningless performance measure: a detector that flags nothing is more than 98 % accurate on every dataset in @fig-dataset-prevalence yet catches no fraud. Evaluation must instead concentrate on the minority class [@he2009learning_imbalanced_data]. Moreover, the tension between false positives and the recall of true positives is at the heart of fraud detection: investigating or blocking a transaction is costly and burdens legitimate customers, so it is not feasible to act on every alarm [@leborgne2022fraud_detection_handbook]. A useful detector must therefore catch as much fraud as possible while keeping its false-positive rate within an operationally acceptable budget.

## From trees to tabular foundation models {#sec-intro-tfm}

Gradient-boosted decision trees (GBDTs) have long been the gold standard for tabular classification, consistently outperforming deep-learning approaches on tabular data [@grinsztajn2022why_tree_based]. Recently, tabular foundation models (TFMs) have become competitive with tuned GBDTs. A TFM is a neural network pretrained on massive amounts of synthetic tabular data. At prediction time it is not trained on the target task at all. The labelled rows are supplied as context, and the model predicts the labels of unseen rows in a single forward pass by conditioning on that context by a mechanism known as in-context learning [@hollmann2025accurate_tabular_foundation_model; @qu2026tabiclv2].

All the information the model uses for prediction must fit into its context window. When the training data exceeds the context budget, some of it must be left out. When working under heavy class imbalance, the way those limited context slots are filled is a crucial modelling decision. 

## Wasting the context window {#sec-context-waste}

The limited context window is currently the principal weakness of TFMs [@hollmann2025accurate_tabular_foundation_model]. When the training data exceed the context budget, standard practice is to fill the window with a uniform random subsample of the training rows. Under heavy class imbalance the cost of this choice is easy to quantify: the expected number of fraud rows in a random context of $B$ rows drawn from data with fraud prevalence $\pi$ is $\pi B$. If the $\pi = 0.17\%$ prevalence of the EU credit-card dataset [@dalpozzolo2015calibrating], and having $B = 10,000$ row context is therefore expected to contain about $\pi B = 17$ fraudulent transactions. When the $\pi$ would be set to closer to real-world fraud rates reported by the EBA and the ECB, the same window would be expected to contain between `r 10000*eba_range[1]` and `r 10000*eba_range[2]` fraudulent transactions [@ebaecb2025_payment_fraud_report], and many random subsamples would contain none at all.

A randomly filled context thus suffers from two distinct deficiencies. First, most of its slots are spent on rows of either class that carry little information about the particular transactions being scored. Second, the few fraud rows that do enter the context form too small a sample for the model to characterise fraudulent behaviour from. These two deficiencies, irrelevance and class imbalance, structure the experimental design of this thesis: each of the four context-construction conditions compared in the experiments targets one of them.

## Retrieval-Augmented Prediction {#sec-intro-rap}

The proposed method of this thesis is Retrieval-Augmented Prediction (RAP), a context-construction strategy. This targets both deficiencies at test time while leaving the pretrained model frozen with no fine-tuning is involved [@thomas2024localpfn_retrieval_finetuning]. It operates through two levers that were discussed as the deficiencies of @sec-context-waste. The first lever is relevance. Instead of a random subsample, nearest-neighbour retrieval fills the context with training rows that are structurally similar to the test transactions being scored, so that no slots are wasted on unrelated observations. The second lever is class balance, the fraud share of the retrieved context is raised to a controlled ratio by oversampling fraud rows. @fig-rap-overview summarises the pipeline. This method will be further discussed in @sec-methods-rap.

![The RAP pipeline. The scored test set is clustered into G groups, and each group retrieves its own context of nearby training rows, with the fraud share raised by oversampling. The labelled context goes through the forward pass together with the group's unlabelled test rows. Popular groups reuse their context across several prediction chunks, and the predicted rows of all passes are collected into one predicted test set.](figures/rap-pipeline.svg){#fig-rap-overview}

The hypothesis under test is that a context that is both relevant and sufficiently populated with the minority class sharpens the posterior predictive distribution the model computes, and thereby improves its ranking of fraudulent transactions [@saerens2002adjusting_priors]. This is a hypothesis rather than a guarantee: the theoretical arguments for localising the context are tied to the synthetic prior on which the models were pretrained and to pretraining-scale context sizes, so any benefit outside that design regime is an empirical question [@nagler2023statistical_foundations_pfn]. The experiments of this thesis are designed to answer it.

## Contribution

The central contribution made in this thesis is a controlled decomposition of the two channels of test-time context manipulation identified in @sec-context-waste. The relevance of the context to the transactions being scored, and its minority-class ratio. Prior work has manipulated one channel at a time either retrieval-based localisation of the context [@thomas2024localpfn_retrieval_finetuning] or rebalancing of its label distribution [@tanna2026data_presentation_resampling]. Whereas the experimental design of this thesis isolates each lever separately and then tests their combination, on frozen tabular foundation models and across datasets whose fraud prevalence spans the range typical of public fraud benchmarks.

The empirical answer to that decomposition is asymmetric. Raising the in-context fraud ratio produces consistent and often large gains, while nearest-neighbour retrieval fails to improve performance and in most configurations degrades it. The full RAP combination is consequently worse than plain rebalancing on every dataset studied. Because retrieval-based context localisation has both theoretical motivation of @nagler2023statistical_foundations_pfn and reported empirical support from @thomas2024localpfn_retrieval_finetuning. Thus this negative result is itself a contribution to the literature. The results are replicated across datasets and traced to a per-dataset mechanism in the experiments. It marks a regime in which the published localisation hypothesis does not transfer. I will be left to further research if these findings generalise to larger transaction datasets.

The positive finding also extends prior work. @tanna2026data_presentation_resampling report that a balanced context improves tabular foundation models on imbalanced credit data at fraud rates of roughly 8–22 %. The experiments here confirm the effect at prevalences one to two orders of magnitude lower (0.13–1.5 %) and evaluate it with the threshold-free ranking metrics that fraud detection requires but that study does not report. The resulting rebalanced-context configuration is, without any per-dataset hyperparameter tuning, competitive with methodologically sound published state-of-the-art results on the benchmark datasets used in this thesis.

## Research questions {#sec-intro-rqs}

The decomposition above translates into four research questions.

**RQ1.** How do tabular foundation models, restricted to a randomly subsampled context, compare with gradient-boosted decision trees trained on the full training data?

**RQ2.** Does controlling the fraud ratio of the context improve foundation-model performance over the natural ratio, and if so, is there an optimal ratio?

**RQ3.** Does filling the context with retrieved rows that are structurally similar to the scored transactions improve performance over a random context?

**RQ4.** How do the two channels interact, does combining retrieval with ratio control add anything beyond either lever alone?

RQ1 establishes the baseline comparison against the dominant model family for tabular data. RQ2 and RQ3 evaluate the two levers of @sec-intro-rap in isolation, and RQ4 evaluates their combination the full RAP method.

## Outline of the thesis {#sec-intro-outline}

The remainder of the thesis is organised as follows. @sec-background introduces the concepts the argument rests on: learning under class imbalance, evaluation with ranking metrics, gradient-boosted trees and the in-context-learning mechanism of tabular foundation models. @sec-related-work reviews prior work on context construction for these models and locates the gap this thesis fills. @sec-data describes the benchmark datasets, their provenance and their limitations. @sec-methods specifies the four context-construction conditions, the RAP pipeline, the models and the evaluation protocol. @sec-results reports the experimental results, one section per research question, and closes with a comparison against published state-of-the-art results. @sec-discussion interprets the findings, states the limitations of the study and outlines future work, and @sec-conclusion summarises the answers to the research questions. The appendices carry supporting material, including a dissection of a degenerate benchmark dataset (@sec-app-cc) and the robustness ablations referenced throughout.

