---
bibliography: ../thesis-refs.bib
---

# Results {#sec-results}

In this chapter we present the results of the tests that were outlined in @sec-methods. At first we will provide the results to the research questions that were expanded in @sec-intro-rqs. Then we will provide additional information on design-space sweeps and ablation studies that are further elaborated in appendixes. Finally, we will compare the best case scenarios of the TFMs to the current state of the art GBDT models from literature.

```{r}
#| label: results-setup
#| include: false

runs <- list.files("results/runs", pattern = "\\.parquet$", full.names = TRUE) |>
  lapply(\(f) select(read_parquet(f), dataset, model, condition, seed,
                     pipeline_version, pr_auc, n_train,
                     context_size, context_fraud_n)) |>
  bind_rows()

formal_datasets <- c("eu_cc", "banksim", "paysim", "fifar", "baf")
gbdt_models <- c("xgboost", "catboost")
v4 <- runs |> filter(pipeline_version == 4, dataset %in% formal_datasets)
c1 <- v4 |> filter(condition == "C1")

## RQ1: family means over the five datasets (per-model means first, so
## the 10-seed models do not outweigh the 3-seed ones inside a family)
fam_means <- c1 |>
  mutate(family = if_else(model %in% gbdt_models, "GBDT", "TFM")) |>
  group_by(family, dataset, model) |>
  summarise(pr = mean(pr_auc),.groups = "drop_last") |>
  summarise(pr = mean(pr),.groups = "drop_last") |>
  summarise(pr = mean(pr),.groups = "drop_last") |> 
  bind_rows( c1 |> 
    filter(model %in% c("tabiclv2", "tabpfn_3")) |> 
    mutate(family = "SOTA TFM") |>
  group_by(family, dataset, model) |>
  summarise(pr = mean(pr_auc),.groups = "drop_last") |>
  summarise(pr = mean(pr),.groups = "drop_last") |>
  summarise(pr = mean(pr),.groups = "drop_last")
  )



fam_of <- \(f) sprintf("%.3f", fam_means$pr[fam_means$family == f])
ntrain_rng <- range(c1$n_train)

## RQ2: paired C1-to-C2 deltas per TFM model and per dataset
paired <- v4 |>
  filter(condition %in% c("C1", "C2"), !model %in% gbdt_models) |>
  select(dataset, model, seed, condition, pr_auc) |>
  tidyr::pivot_wider(names_from = condition, values_from = pr_auc) |>
  filter(!is.na(C1), !is.na(C2)) |>
  mutate(delta = C2 - C1)
gain_by_model <- paired |> group_by(model) |> summarise(d = mean(delta))
gain_of <- \(m) sprintf("%+.3f", gain_by_model$d[gain_by_model$model == m])
gain_paysim <- paired |>
  filter(dataset == "paysim") |> summarise(d = mean(delta)) |> pull(d)
canon <- paired |> filter(dataset == "paysim", model == "tabpfn_v2") |>
  summarise(C1 = mean(C1), C2 = mean(C2))
canon_fr <- v4 |>
  filter(dataset == "paysim", model == "tabpfn_v2",
         condition %in% c("C1", "C2")) |>
  group_by(condition) |>
  summarise(fr = mean(context_fraud_n), .groups = "drop")
canon_fr_of <- \(cond) comma(round(canon_fr$fr[canon_fr$condition == cond]))

## realized in-context fraud shares
c1_share <- c1 |> filter(!model %in% gbdt_models) |>
  summarise(s = mean(context_fraud_n / context_size)) |> pull(s)
c2_share <- v4 |>
  filter(condition == "C2", !model %in% gbdt_models, model != "tabpfn_v2") |>
  group_by(dataset) |>
  summarise(s = mean(context_fraud_n / context_size), .groups = "drop")
c2_share_eucc <- c2_share$s[c2_share$dataset == "eu_cc"]
c2_share_rng <- range(c2_share$s[c2_share$dataset != "eu_cc"])

## C2 fraud-ratio sweep on eu_cc (50k context) and the XGBoost reference
c2grid <- list.files("results/runs_c2grid", pattern = "\\.parquet$",
                     full.names = TRUE) |>
  lapply(read_parquet) |>
  bind_rows()
sweep_datasets <- c("eu_cc", "banksim", "paysim","baf")
sweep_labels <- c("EU CC", "BankSim", "PaySim", "BAF")
rq2_summ <- c2grid |>
  filter(context_size_req == 50000) |>
  group_by(dataset, model, fraud_ratio) |>
  summarise(mean_pr = mean(pr_auc), sd_pr = sd(pr_auc), .groups = "drop") |>
  mutate(model = factor(model, c("tabpfn_3", "tabiclv2"),
                        c("TabPFN v3", "TabICLv2")),
         dataset = factor(dataset, sweep_datasets, sweep_labels))
xgb_ref <- c1 |>
  filter(model == "xgboost", dataset %in% sweep_datasets) |>
  group_by(dataset) |>
  summarise(pr = mean(pr_auc), .groups = "drop") |>
  mutate(dataset = factor(dataset, sweep_datasets, sweep_labels))
```

## Research question 1: GBDTs vs TFMs on randomly sampled context windows {#sec-results-rq1}

Under a random context, context-limited TFMs and full-data trees are roughly on par. Averaged over the five formal datasets, the GBDT family reaches a mean PR-AUC of `r fam_of("GBDT")` against `r fam_of("TFM")` for the whole TFM family and the mean of the two SOTA TFM models it climbs to `{r} fam_of("SOTA TFM")`. Comparing the best-in-family TFM to the best GBDT on each dataset gives two TFM wins (eu_cc and banksim), one effective tie (fifar) and two losses (paysim and baf). @fig-rq1-c1-prauc shows the per-model breakdown.

```{r}
#| label: fig-rq1-c1-prauc
#| fig-cap: "PR-AUC under condition C1 (random context) on the five formal-grid datasets. Columns are means over the seeds: three for the other models, ten for TabPFN v3 and TabICLv2. Whiskers extend one standard deviation above the mean; FiFAR has none because it runs on a single pre-defined split. Striped columns mark the GBDT baselines, which are trained on the full training split."
#| fig-width: 6.5
#| fig-height: 3.8
#| warning: false

c1_summ <- c1 |>
  group_by(dataset, model) |>
  summarise(mean_pr = mean(pr_auc), sd_pr = sd(pr_auc), .groups = "drop") |>
  mutate(
    family = if_else(model %in% c("xgboost", "catboost"), "GBDT", "TFM"),
    model = factor(
      model,
      c("xgboost", "catboost", "tabpfn_v2", "tabpfn_25",
        "tabpfn_26", "tabpfn_3", "tabiclv2"),
      c("XGBoost", "CatBoost", "TabPFN v2", "TabPFN 2.5",
        "TabPFN 2.6", "TabPFN v3", "TabICLv2")
    ),
    dataset = factor(
      dataset,
      c("eu_cc", "banksim", "paysim", "fifar", "baf"),
      c("EU CC", "BankSim", "PaySim", "FiFAR", "BAF")
    )
  )

ggplot(c1_summ, aes(dataset, mean_pr, fill = model)) +
  geom_col_pattern(
    aes(pattern = family),
    position = position_dodge2(padding = 0.15), width = 0.85,
    colour = "grey25", linewidth = 0.15,
    pattern_colour = NA, pattern_fill = "white",
    pattern_angle = 45, pattern_density = 0.25, pattern_spacing = 0.02,
    pattern_key_scale_factor = 0.6
  ) +
  geom_errorbar(aes(ymin = mean_pr, ymax = mean_pr + sd_pr),
                position = position_dodge2(padding = 0.15),
                width = 0.85, linewidth = 0.35, colour = "grey25",
                na.rm = TRUE) +
  scale_fill_manual(
    values = c("#2a78d6", "#008300", "#e87ba4", "#eda100",
               "#1baf7a", "#eb6834", "#4a3aa7"),
    name = NULL
  ) +
  scale_pattern_manual(values = c(GBDT = "stripe", TFM = "none")) +
  scale_y_continuous(limits = c(0, 1.1), expand = expansion(0)) +
  labs(x = NULL, y = "PR-AUC") +
  guides(
    pattern = "none",
    fill = guide_legend(
      nrow = 2, byrow = TRUE,
      override.aes = list(
        pattern = c("stripe", "stripe", "none", "none", "none", "none", "none")
      )
    )
  ) +
  theme_minimal(base_size = 10) +
  theme(
    legend.position = "bottom",
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    axis.ticks.x = element_blank()
  )
```

The comparison is intentionally asymmetric. The GBDTs are trained on the full training split, from `r comma(ntrain_rng[1])` rows on eu_cc to `r comma(ntrain_rng[2])` on paysim, while the TFMs see only a random subsample of at most 50,000 of those rows (@sec-methods-conditions). The asymmetry persists under every later condition, because no context-construction strategy adds data. The balancing gains of @sec-results-rq2 are achieved despite it. The answer to RQ1 is that the TFMs are at par with the GBDT baselines even in the worst-case scenario of random subsampling, as can be seen in @fig-rq1-c1-prauc.

## Research question 2: the fraud ratio's effect on prediction performance {#sec-results-rq2}

The standalone deficiency of a random context is its in-context fraud count. A C1 context holds the natural fraud rate whatever its size, on average `r sprintf("%.1f", 100 * c1_share)` % across the runs. So even a 50,000-row window carries only a few hundred fraud rows, and the 10,000-row window of TabPFN v2 can carry only a handful. Where this deficiency is worst, the gains from balancing are highest. The mean paired C1-to-C2 lift in PR-AUC is largest for the smallest-context model, TabPFN v2, at `r gain_of("tabpfn_v2")`, against `r gain_of("tabpfn_25")` for TabPFN 2.5, `r gain_of("tabpfn_26")` for TabPFN 2.6, `r gain_of("tabpfn_3")` for TabPFN v3 and `r gain_of("tabiclv2")` for TabICLv2. Across datasets the lift is largest on the most imbalanced dataset, paysim, at `r sprintf("%+.3f", gain_paysim)`. The canonical case combines both: on paysim, TabPFN v2 holds on average `r canon_fr_of("C1")` fraud rows in its C1 window and scores a PR-AUC of `r sprintf("%.3f", canon$C1)`; balancing lifts the fraud count to `r canon_fr_of("C2")` and the PR-AUC to `r sprintf("%.3f", canon$C2)`.

@fig-rq2-ratio-sweep traces the ratio lever directly on the three datasets covered by the ratio sweep which is delved depper in  @sec-app-c2-sweep. Results are for the two strongest TFMs at a fixed 50,000-row stratified random context. BAF and Fifar sit at the separability floor and were not  thus part of this additional ratio and context size sweep. The optimal ratio is dataset-dependent but broad. On eu_cc and banksim, PR-AUC varies little for ratios up to about 0.10 and declines once the ratio reaches 0.20. On eu_cc the high ratios additionally imply heavy duplication of its roughly 394 <!-- TODO: inline R instead of you running it yourself --> unique training fraud rows. On paysim, PR-AUC rises up to a ratio of 0.30 and stays roughly flat thereafter. Ratios near 0.10 sit close to the peak on every dataset.  The models sit at or above the full-data XGBoost baseline through the middle of the sweep on all three datasets, and fall below it only at the extremes.

```{r}
#| label: fig-rq2-ratio-sweep
#| fig-cap: "PR-AUC against the requested in-context fraud ratio, for the two strongest TFMs at a fixed 50,000-row stratified random context. Lines are means over three seeds; the bands span one standard deviation on each side of the mean. Note the free y-scales. The dotted lines mark the per-dataset seed-mean of the full-data XGBoost baseline under C1."
#| fig-width: 6.5
#| fig-height: 8
#| warning: false

ggplot(rq2_summ, aes(fraud_ratio, mean_pr, colour = model, fill = model)) +
  geom_hline(data = xgb_ref, aes(yintercept = pr),
             linetype = "dotted", colour = "grey25") +
  geom_ribbon(aes(ymin = mean_pr - sd_pr, ymax = mean_pr + sd_pr),
              alpha = 0.2, colour = NA) +
  geom_line(linewidth = 0.6) +
  geom_point(size = 1.1) +
  facet_wrap(~ dataset, nrow = 4, scales = "free_y") +
  scale_colour_manual(values = c("TabPFN v3" = "#eb6834",
                                 "TabICLv2" = "#4a3aa7"), name = NULL) +
  scale_fill_manual(values = c("TabPFN v3" = "#eb6834",
                               "TabICLv2" = "#4a3aa7"), name = NULL) +
  scale_x_continuous(breaks = c(0.02, 0.1, 0.2, 0.3, 0.4)) +
  labs(x = "Requested in-context fraud ratio", y = "PR-AUC") +
  theme_minimal(base_size = 10) +
  theme(legend.position = "bottom", panel.grid.minor = element_blank())
```

The C1-to-C2 gain is measured on threshold-free ranking metrics. On our interpretive reading of the in-context versus in-weight distinction, stated in @sec-bg-eval, such a gain reflects genuine discrimination rather than recalibration. As a post-hoc prior correction is a monotone rescaling of a fixed scorer's outputs and cannot move a ranking metric [@saerens2002adjusting_priors]. 

What is meant here by balanced requires a bit of further clarificationcare. C2 reaches a true 50/50 context only on TabPFN v2, whose 10,000-row window is small enough for the available fraud rows to fill half of it. On the models with 48,000 to 50,000-row budgets, which produce the strongest results, the realized fraud share reaches only `r round(100 * c2_share_rng[1])` % to `r round(100 * c2_share_rng[2])` % on four datasets and `r sprintf("%.1f", 100 * c2_share_eucc)` % on eu_cc. The sweep visualised in @fig-rq2-ratio-sweep and further investigated in @sec-app-c2-sweep suggests that pushing the realized ratio higher would have brought little value. On eu_cc and banksim performance degrades at ratios of 0.20 and above, and on paysim the gains past 0.10 are marginal. The effective lever is therefore to include enough fraud tags, on the order of 5 % to 10 % of the context, while avoiding heavy duplication.

## Research question 3: the relevance of Relevance {#sec-results-rq3}

```{r}
#| label: rq3-setup
#| include: false

rq3_md_lv <- c("tabpfn_3", "tabiclv2")
rq3_md_lab <- c("TabPFN v3", "TabICLv2")
rq3_cols <- c("TabPFN v3" = "#eb6834", "TabICLv2" = "#4a3aa7")
rq3_shapes <- c("TabPFN v3" = 16, "TabICLv2" = 17)

## C3 vs C1 on matched seeds. Needs the C3-only grouping columns, which
## results-setup does not carry; pre-Phase-3 C1/C2 parquets lack them, so
## select per file and let bind_rows() NA-fill.
rq3_cols_sel <- c("dataset", "model", "condition", "seed", "pipeline_version",
                  "n_estimators", "pr_auc", "n_groups", "group_silhouette")
rq3_cmp <- list.files("results/runs", pattern = "\\.parquet$",
                      full.names = TRUE) |>
  lapply(\(f) { d <- read_parquet(f)
                d[, intersect(rq3_cols_sel, names(d)), drop = FALSE] }) |>
  bind_rows() |>
  filter(pipeline_version == 4, n_estimators == 4,
         dataset %in% formal_datasets, model %in% rq3_md_lv,
         condition %in% c("C1", "C3"))
rq3_c3 <- rq3_cmp |> filter(condition == "C3")
rq3_pairs <- rq3_cmp |>
  semi_join(distinct(rq3_c3, dataset, model, seed),
            by = c("dataset", "model", "seed")) |>
  select(dataset, model, seed, condition, pr_auc) |>
  tidyr::pivot_wider(names_from = condition, values_from = pr_auc) |>
  mutate(delta = C3 - C1)
rq3_delta <- rq3_pairs |>
  group_by(dataset, model) |>
  summarise(d = mean(delta), .groups = "drop")
rq3_d_of <- \(ds, m, dg = 2) sprintf(paste0("%+.", dg, "f"),
  rq3_delta$d[rq3_delta$dataset == ds & rq3_delta$model == m])
rq3_c3_of <- \(ds, f) sprintf("%.2f", f(rq3_c3$pr_auc[rq3_c3$dataset == ds]))

## silhouette-chosen grouping under C3
rq3_sil <- rq3_c3 |>
  group_by(dataset) |>
  summarise(G = round(mean(n_groups)),
            sil = mean(group_silhouette), .groups = "drop")
rq3_G_of <- \(ds) rq3_sil$G[rq3_sil$dataset == ds]
rq3_sil_of <- \(ds) sprintf("%.2f", rq3_sil$sil[rq3_sil$dataset == ds])

## natural-ratio arms of the relevance ablation (G = 32, 50k context)
rq3_abl_nat <- list.files("results/runs_ablation", pattern = "\\.parquet$",
                          full.names = TRUE) |>
  lapply(read_parquet) |>
  bind_rows() |>
  filter(ratio_spec %in% "natural", G == 32) |>
  group_by(dataset, arm) |>
  summarise(pr = mean(pr_auc), .groups = "drop")
rq3_abl_of <- \(ds, a) sprintf("%.3f",
  rq3_abl_nat$pr[rq3_abl_nat$dataset == ds & rq3_abl_nat$arm == a])

## design-space sweep: PR-AUC and legit score by forced group count
rq3_rapds <- list.files("results/runs_rapds", pattern = "\\.parquet$",
                        full.names = TRUE) |>
  lapply(read_parquet) |>
  bind_rows()

rq3_g <- rq3_rapds |>
  group_by(dataset, model, granularity) |>
  summarise(`PR-AUC` = mean(pr_auc),
            `Mean legitimate score` = mean(legit_score), .groups = "drop") |>
  tidyr::pivot_longer(c(`PR-AUC`, `Mean legitimate score`),
                      names_to = "measure", values_to = "value") |>
  mutate(dataset = factor(dataset, c("eu_cc", "banksim", "paysim"),
                          c("EU CC", "BankSim", "PaySim")),
         model = factor(model, rq3_md_lv, rq3_md_lab),
         measure = factor(measure, c("PR-AUC", "Mean legitimate score")))
         
rq3_refs <- v4 |>
  filter(condition %in% c("C1", "C2"), model %in% rq3_md_lv,
         seed %in% c(7, 42, 123),
         dataset %in% c("eu_cc", "banksim", "paysim")) |>
  group_by(dataset, condition) |>
  summarise(pr = mean(pr_auc), .groups = "drop") |>
  mutate(dataset = factor(dataset, c("eu_cc", "banksim", "paysim"),
                          c("EU CC", "BankSim", "PaySim")),
         measure = factor("PR-AUC", c("PR-AUC", "Mean legitimate score")))
```

The primary finding to the third research question is negative. Filling the context by kNN retrieval at the natural fraud ratio (C3) scores worse than the random context (C1) on four of the five datasets, sometimes with a catastrophic fall on the performance. The mean PR-AUC delta is for TabPFN v3  `r rq3_d_of("paysim", "tabpfn_3")` on paysim and `r rq3_d_of("eu_cc", "tabpfn_3")` on eu_cc, and respectively for TabICL v2 `r rq3_d_of("paysim", "tabiclv2")` and `r rq3_d_of("eu_cc", "tabiclv2")`. The banksim drop is mild for TabPFN v3 `r rq3_d_of("banksim", "tabpfn_3")` and for TabICL 2 `r rq3_d_of("banksim", "tabiclv2")`, and the fifar contrast of `r rq3_d_of("fifar", "tabpfn_3")` and `r rq3_d_of("fifar", "tabiclv2")` rests on the single pre-defined split. All contrasts in this section use only the same seeded runs experiment C3 ran. Experiment C1 carried ten seeds for these two models shown on @fig-rq1-c1-prauc, so unmatched means would mix the seed sets. On baf the outcome is a tie with deltas of `r rq3_d_of("baf", "tabpfn_3", 3)` and `r rq3_d_of("baf", "tabiclv2", 3)`.

The per-seed spread is itself a finding. On eu_cc the worst seed falls to a PR-AUC of `r rq3_c3_of("eu_cc", min)` while the best stays close to C1 at `r rq3_c3_of("eu_cc", max)`. On paysim every seed collapses, from `r rq3_c3_of("paysim", max)` down to `r rq3_c3_of("paysim", min)`.

A drop under C3 does not by itself isolate relevance, because the realized fraud share of a retrieved context drifts with the local neighbourhood and the silhouette criterion chooses the grouping (@sec-methods-conditions). The ratio-matched relevance ablation of @sec-app-relevance-ablation therefore carries the verdict on RQ3 it holds the grouping fixed at G=32 and enforces the fraud ratio, so only the selection mechanism varies. 

The grouping the silhouette criterion picks under C3 is coarse: on average `r rq3_G_of("banksim")` groups on banksim, `r rq3_G_of("paysim")` on paysim and `r rq3_G_of("fifar")` on fifar, `r rq3_G_of("eu_cc")` on eu_cc and `r rq3_G_of("baf")` on baf. Cluster quality is weak outside banksim, with mean silhouette scores of `r rq3_sil_of("banksim")` on banksim against `r rq3_sil_of("paysim")` on paysim, `r rq3_sil_of("eu_cc")` on eu_cc, `r rq3_sil_of("fifar")` on fifar and `r rq3_sil_of("baf")` on baf. Coarseness alone does not explain the loss, however. We can see in  @fig-rq3-granularity that forcing finer groupings in the design-space sweep recovers part of it but asymptotes below C2 on all six dataset and model pairs.


```{r}
#| label: fig-rq3-granularity
#| fig-cap: "The Relevance sweep under specified group counts, means pooled over metrics, ratios and seeds. Top row has PR-AUC recovers with finer G on every dataset but stays below the matched-seed C2 mean (dashed) on all six dataset and model pairs on BankSim the C1 and C2 references coincide. Bottom row has the mean predicted score on legitimate (non-fraud) rows falls with finer G everywhere, so retrieved-context legit inflation vanishes at fine G on EU CC and BankSim while PaySim remains collapsed, ruling that mechanism out as the PaySim explanation."
#| fig-width: 6.5
#| fig-height: 4.6
#| warning: false

ggplot(rq3_g, aes(granularity, value, colour = model, shape = model)) +
  geom_hline(data = rq3_refs, aes(yintercept = pr, linetype = condition),
             colour = "#52514e", linewidth = 0.4) +
  geom_line(aes(group = model), linewidth = 0.6) +
  geom_point(size = 1.8) +
  facet_grid(measure ~ dataset, scales = "free_y", switch = "y") +
  scale_colour_manual(values = rq3_cols, name = NULL) +
  scale_shape_manual(values = rq3_shapes, name = NULL) +
  scale_linetype_manual(values = c(C1 = "dotted", C2 = "dashed"), name = NULL) +
  scale_x_continuous(trans = "log2", breaks = c(8, 16, 32, 64)) +
  labs(x = "Forced group count G (log scale)", y = NULL) +
  theme_minimal(base_size = 10) +
  theme(legend.position = "bottom", panel.grid.minor = element_blank(),
        strip.placement = "outside")
```

The ablation and the design-space sweep cover eu_cc, banksim and paysim only, so for baf and fifar the evidence is limited to the contrasts of the main experiments. The per-dataset answer to RQ3 is thus negative but graded: retrieval-relevance is strongly harmful on paysim, mildly harmful on banksim, neutral on eu_cc once the ratio and the grouping are controlled, a tie at the floor on baf, and directional only on fifar. This is a regime and domain transfer gap rather than a general refutation of retrieval. Frozen kNN retrieval helps on balanced general tabular data [@thomas2024localpfn_retrieval_finetuning], while under fraud-grade imbalance our result confirms the strong-random-baseline prior [@feuer2023scaling_tabpfn] and that naïve distance-based context construction methods can affect performance [@pham2026aware].

