---
bibliography: ../thesis-refs.bib
---

# Results {#sec-results}

In this chapter we present the results of the tests that were outlined in @sec-methods. At first we will provide the results to the research questions that were expanded in @sec-intro-rqs. Then we will provide additional information on design-space sweeps and ablation studies that are further elaborated in appendixes. Finally, we will compare the best case scenarios of the TFMs to the current state of the art GBDT models from literature.

```{r}
#| label: results-setup
#| include: false
library(dplyr)
library(ggplot2)
library(scales)
library(ggpattern)
library(gt)

runs <- list.files("results/runs", pattern = "\\.parquet$", full.names = TRUE) |>
  lapply(\(f) select(nanoparquet::read_parquet(f), dataset, model, condition, seed,
                     pipeline_version, pr_auc, n_train, n_fraud_train,
                     context_size, context_fraud_n)) |>
  bind_rows()

formal_datasets <- c("eu_cc", "banksim", "paysim", "fifar", "baf")
gbdt_models <- c("xgboost", "catboost")
v4 <- runs |> filter(pipeline_version == 4, dataset %in% formal_datasets)
c1 <- v4 |> filter(condition == "C1")

## RQ1: family means over the five datasets (per-model means first, so
## the 10-seed models do not outweigh the 3-seed ones inside a family)
fam_means <- c1 |>
  mutate(family = if_else(model %in% gbdt_models, "GBDT", "TFM")) |>
  group_by(family, dataset, model) |>
  summarise(pr = mean(pr_auc),.groups = "drop_last") |>
  summarise(pr = mean(pr),.groups = "drop_last") |>
  summarise(pr = mean(pr),.groups = "drop_last") |> 
  bind_rows( c1 |> 
    filter(model %in% c("tabiclv2", "tabpfn_3")) |> 
    mutate(family = "SOTA TFM") |>
  group_by(family, dataset, model) |>
  summarise(pr = mean(pr_auc),.groups = "drop_last") |>
  summarise(pr = mean(pr),.groups = "drop_last") |>
  summarise(pr = mean(pr),.groups = "drop_last")
  )



fam_of <- \(f) sprintf("%.3f", fam_means$pr[fam_means$family == f])
ntrain_rng <- range(c1$n_train)

## RQ2: paired C1-to-C2 deltas per TFM model and per dataset
paired <- v4 |>
  filter(condition %in% c("C1", "C2"), !model %in% gbdt_models) |>
  select(dataset, model, seed, condition, pr_auc) |>
  tidyr::pivot_wider(names_from = condition, values_from = pr_auc) |>
  filter(!is.na(C1), !is.na(C2)) |>
  mutate(delta = C2 - C1)
gain_by_model <- paired |> group_by(model) |> summarise(d = mean(delta))
gain_of <- \(m) sprintf("%+.3f", gain_by_model$d[gain_by_model$model == m])
gain_paysim <- paired |>
  filter(dataset == "paysim") |> summarise(d = mean(delta)) |> pull(d)
canon <- paired |> filter(dataset == "paysim", model == "tabpfn_v2") |>
  summarise(C1 = mean(C1), C2 = mean(C2))
canon_fr <- v4 |>
  filter(dataset == "paysim", model == "tabpfn_v2",
         condition %in% c("C1", "C2")) |>
  group_by(condition) |>
  summarise(fr = mean(context_fraud_n), .groups = "drop")
canon_fr_of <- \(cond) comma(round(canon_fr$fr[canon_fr$condition == cond]))

## realized in-context fraud shares
c1_share <- c1 |> filter(!model %in% gbdt_models) |>
  summarise(s = mean(context_fraud_n / context_size)) |> pull(s)
c2_share <- v4 |>
  filter(condition == "C2", !model %in% gbdt_models, model != "tabpfn_v2") |>
  group_by(dataset) |>
  summarise(s = mean(context_fraud_n / context_size), .groups = "drop")
c2_share_eucc <- c2_share$s[c2_share$dataset == "eu_cc"]
c2_share_rng <- range(c2_share$s[c2_share$dataset != "eu_cc"])

## C2 fraud-ratio sweep on eu_cc (50k context) and the XGBoost reference
c2grid <- list.files("results/runs_c2grid", pattern = "\\.parquet$",
                     full.names = TRUE) |>
  lapply(nanoparquet::read_parquet) |>
  bind_rows()
## only these three datasets carry the full ratio sweep at the 50k context;
## baf and fifar sit in runs_c2grid at ratio 0.10 only and are excluded
sweep_datasets <- c("eu_cc", "banksim", "paysim")
sweep_labels <- c("EU CC", "BankSim", "PaySim")
rq2_summ <- c2grid |>
  filter(context_size_req == 50000, dataset %in% sweep_datasets) |>
  group_by(dataset, model, fraud_ratio) |>
  summarise(mean_pr = mean(pr_auc), sd_pr = sd(pr_auc), .groups = "drop") |>
  mutate(model = factor(model, c("tabpfn_3", "tabiclv2"),
                        c("TabPFN v3", "TabICLv2")),
         dataset = factor(dataset, sweep_datasets, sweep_labels))
xgb_ref <- c1 |>
  filter(model == "xgboost", dataset %in% sweep_datasets) |>
  group_by(dataset) |>
  summarise(pr = mean(pr_auc), .groups = "drop") |>
  mutate(dataset = factor(dataset, sweep_datasets, sweep_labels))
```

## Research question 1: GBDTs vs TFMs on randomly sampled context windows {#sec-results-rq1}

Under a random context, context-limited TFMs and full-data trees are roughly on par. Averaged over the five formal datasets, the GBDT family reaches a mean PR-AUC of `r fam_of("GBDT")` against `r fam_of("TFM")` for the whole TFM family and the mean of the two SOTA TFM models it climbs to `{r} fam_of("SOTA TFM")`. Comparing the best-in-family TFM to the best GBDT on each dataset gives two TFM wins (eu_cc and banksim), one effective tie (fifar) and two losses (paysim and baf). @fig-rq1-c1-prauc shows the per-model breakdown.

```{r}
#| label: fig-rq1-c1-prauc
#| fig-cap: "PR-AUC under condition C1 (random context) on the five formal-grid datasets. Columns are means over the seeds: three for the other models, ten for TabPFN v3 and TabICLv2. Whiskers extend one standard deviation above the mean; FiFAR has none because it runs on a single pre-defined split. Striped columns mark the GBDT baselines, which are trained on the full training split."
#| fig-width: 6.5
#| fig-height: 3.8
#| warning: false

c1_summ <- c1 |>
  group_by(dataset, model) |>
  summarise(mean_pr = mean(pr_auc), sd_pr = sd(pr_auc), .groups = "drop") |>
  mutate(
    family = if_else(model %in% c("xgboost", "catboost"), "GBDT", "TFM"),
    model = factor(
      model,
      c("xgboost", "catboost", "tabpfn_v2", "tabpfn_25",
        "tabpfn_26", "tabpfn_3", "tabiclv2"),
      c("XGBoost", "CatBoost", "TabPFN v2", "TabPFN 2.5",
        "TabPFN 2.6", "TabPFN v3", "TabICLv2")
    ),
    dataset = factor(
      dataset,
      c("eu_cc", "banksim", "paysim", "fifar", "baf"),
      c("EU CC", "BankSim", "PaySim", "FiFAR", "BAF")
    )
  )

ggplot(c1_summ, aes(dataset, mean_pr, fill = model)) +
  geom_col_pattern(
    aes(pattern = family),
    position = position_dodge2(padding = 0.15), width = 0.85,
    colour = "grey25", linewidth = 0.15,
    pattern_colour = NA, pattern_fill = "white",
    pattern_angle = 45, pattern_density = 0.25, pattern_spacing = 0.02,
    pattern_key_scale_factor = 0.6
  ) +
  geom_errorbar(aes(ymin = mean_pr, ymax = mean_pr + sd_pr),
                position = position_dodge2(padding = 0.15),
                width = 0.85, linewidth = 0.35, colour = "grey25",
                na.rm = TRUE) +
  scale_fill_manual(
    values = c("#2a78d6", "#008300", "#e87ba4", "#eda100",
               "#1baf7a", "#eb6834", "#4a3aa7"),
    name = NULL
  ) +
  scale_pattern_manual(values = c(GBDT = "stripe", TFM = "none")) +
  scale_y_continuous(limits = c(0, 1.1), expand = expansion(0)) +
  labs(x = NULL, y = "PR-AUC") +
  guides(
    pattern = "none",
    fill = guide_legend(
      nrow = 2, byrow = TRUE,
      override.aes = list(
        pattern = c("stripe", "stripe", "none", "none", "none", "none", "none")
      )
    )
  ) +
  theme_minimal(base_size = 10) +
  theme(
    legend.position = "bottom",
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    axis.ticks.x = element_blank()
  )
```

The comparison is intentionally asymmetric. The GBDTs are trained on the full training split, from `r comma(ntrain_rng[1])` rows on eu_cc to `r comma(ntrain_rng[2])` on paysim, while the TFMs see only a random subsample of at most 50,000 of those rows (@sec-methods-conditions). The asymmetry persists under every later condition, because no context-construction strategy adds data. The balancing gains of @sec-results-rq2 are achieved despite it. The answer to RQ1 is that the TFMs are at par with the GBDT baselines even in the worst-case scenario of random subsampling, as can be seen in @fig-rq1-c1-prauc.

## Research question 2: the fraud ratio's effect on prediction performance {#sec-results-rq2}

The standalone deficiency of a random context is its in-context fraud count. A C1 context holds the natural fraud rate whatever its size, on average `r sprintf("%.1f", 100 * c1_share)` % across the runs. So even a 50,000-row window carries only a few hundred fraud rows, and the 10,000-row window of TabPFN v2 can carry only a handful. Where this deficiency is worst, the gains from balancing are highest. The mean paired C1-to-C2 lift in PR-AUC is largest for the smallest-context model, TabPFN v2, at `r gain_of("tabpfn_v2")`, against `r gain_of("tabpfn_25")` for TabPFN 2.5, `r gain_of("tabpfn_26")` for TabPFN 2.6, `r gain_of("tabpfn_3")` for TabPFN v3 and `r gain_of("tabiclv2")` for TabICLv2. Across datasets the lift is largest on the most imbalanced dataset, paysim, at `r sprintf("%+.3f", gain_paysim)`. The canonical case combines both: on paysim, TabPFN v2 holds on average `r canon_fr_of("C1")` fraud rows in its C1 window and scores a PR-AUC of `r sprintf("%.3f", canon$C1)`; balancing lifts the fraud count to `r canon_fr_of("C2")` and the PR-AUC to `r sprintf("%.3f", canon$C2)`.

@fig-rq2-ratio-sweep traces the ratio lever directly on the three datasets covered by the ratio sweep which is delved depper in  @sec-app-c2-sweep. Results are for the two strongest TFMs at a fixed 50,000-row stratified random context. BAF and Fifar sit at the separability floor and were not  thus part of this additional ratio and context size sweep. The optimal ratio is dataset-dependent but broad. On eu_cc and banksim, PR-AUC varies little for ratios up to about 0.10 and declines once the ratio reaches 0.20. On eu_cc the high ratios additionally imply heavy duplication of its roughly `r round(mean(v4$n_fraud_train[v4$dataset == "eu_cc"]))` unique training fraud rows. On paysim, PR-AUC rises up to a ratio of 0.30 and stays roughly flat thereafter. Ratios near 0.10 sit close to the peak on every dataset.  The models sit at or above the full-data XGBoost baseline through the middle of the sweep on all three datasets, and fall below it only at the extremes.

```{r}
#| label: fig-rq2-ratio-sweep
#| fig-cap: "PR-AUC against the requested in-context fraud ratio, for the two strongest TFMs at a fixed 50,000-row stratified random context, on the three datasets that carry the full sweep. Lines are means over three seeds; the bands span one standard deviation on each side of the mean. Note the free y-scales. The dotted lines mark the per-dataset seed-mean of the full-data XGBoost baseline under C1."
#| fig-width: 6.5
#| fig-height: 8
#| warning: false

ggplot(rq2_summ, aes(fraud_ratio, mean_pr, colour = model, fill = model)) +
  geom_hline(data = xgb_ref, aes(yintercept = pr),
             linetype = "dotted", colour = "grey25") +
  geom_ribbon(aes(ymin = mean_pr - sd_pr, ymax = mean_pr + sd_pr),
              alpha = 0.2, colour = NA) +
  geom_line(linewidth = 0.6) +
  geom_point(size = 1.1) +
  facet_wrap(~ dataset, nrow = 3, scales = "free_y") +
  scale_colour_manual(values = c("TabPFN v3" = "#eb6834",
                                 "TabICLv2" = "#4a3aa7"), name = NULL) +
  scale_fill_manual(values = c("TabPFN v3" = "#eb6834",
                               "TabICLv2" = "#4a3aa7"), name = NULL) +
  scale_x_continuous(breaks = c(0.02, 0.1, 0.2, 0.3, 0.4)) +
  labs(x = "Requested in-context fraud ratio", y = "PR-AUC") +
  theme_minimal(base_size = 10) +
  theme(legend.position = "bottom", panel.grid.minor = element_blank())
```

The C1-to-C2 gain is measured on threshold-free ranking metrics. On our interpretive reading of the in-context versus in-weight distinction, stated in @sec-bg-eval, such a gain reflects genuine discrimination rather than recalibration. As a post-hoc prior correction is a monotone rescaling of a fixed scorer's outputs and cannot move a ranking metric [@saerens2002adjusting_priors]. 

What is meant here by balanced requires a bit of further clarificationcare. C2 reaches a true 50/50 context only on TabPFN v2, whose 10,000-row window is small enough for the available fraud rows to fill half of it. On the models with 48,000 to 50,000-row budgets, which produce the strongest results, the realized fraud share reaches only `r round(100 * c2_share_rng[1])` % to `r round(100 * c2_share_rng[2])` % on four datasets and `r sprintf("%.1f", 100 * c2_share_eucc)` % on eu_cc. The sweep visualised in @fig-rq2-ratio-sweep and further investigated in @sec-app-c2-sweep suggests that pushing the realized ratio higher would have brought little value. On eu_cc and banksim performance degrades at ratios of 0.20 and above, and on paysim the gains past 0.10 are marginal. The effective lever is therefore to include enough fraud tags, on the order of 5 % to 10 % of the context, while avoiding heavy duplication.

## Research question 3: the relevance of Relevance {#sec-results-rq3}

```{r}
#| label: rq3-setup
#| include: false

rq3_md_lv <- c("tabpfn_3", "tabiclv2")
rq3_md_lab <- c("TabPFN v3", "TabICLv2")
rq3_cols <- c("TabPFN v3" = "#eb6834", "TabICLv2" = "#4a3aa7")
rq3_shapes <- c("TabPFN v3" = 16, "TabICLv2" = 17)

## C3 vs C1 on matched seeds. Needs the C3-only grouping columns, which
## results-setup does not carry; pre-Phase-3 C1/C2 parquets lack them, so
## select per file and let bind_rows() NA-fill.
rq3_cols_sel <- c("dataset", "model", "condition", "seed", "pipeline_version",
                  "n_estimators", "pr_auc", "n_groups", "group_silhouette")
rq3_cmp <- list.files("results/runs", pattern = "\\.parquet$",
                      full.names = TRUE) |>
  lapply(\(f) { d <- nanoparquet::read_parquet(f)
                d[, intersect(rq3_cols_sel, names(d)), drop = FALSE] }) |>
  bind_rows() |>
  filter(pipeline_version == 4, n_estimators == 4,
         dataset %in% formal_datasets, model %in% rq3_md_lv,
         condition %in% c("C1", "C3"))
rq3_c3 <- rq3_cmp |> filter(condition == "C3")
rq3_pairs <- rq3_cmp |>
  semi_join(distinct(rq3_c3, dataset, model, seed),
            by = c("dataset", "model", "seed")) |>
  select(dataset, model, seed, condition, pr_auc) |>
  tidyr::pivot_wider(names_from = condition, values_from = pr_auc) |>
  mutate(delta = C3 - C1)
rq3_delta <- rq3_pairs |>
  group_by(dataset, model) |>
  summarise(d = mean(delta), .groups = "drop")
rq3_d_of <- \(ds, m, dg = 2) sprintf(paste0("%+.", dg, "f"),
  rq3_delta$d[rq3_delta$dataset == ds & rq3_delta$model == m])
rq3_c3_of <- \(ds, f) sprintf("%.2f", f(rq3_c3$pr_auc[rq3_c3$dataset == ds]))

## silhouette-chosen grouping under C3
rq3_sil <- rq3_c3 |>
  group_by(dataset) |>
  summarise(G = round(mean(n_groups)),
            sil = mean(group_silhouette), .groups = "drop")
rq3_G_of <- \(ds) rq3_sil$G[rq3_sil$dataset == ds]
rq3_sil_of <- \(ds) sprintf("%.2f", rq3_sil$sil[rq3_sil$dataset == ds])

## natural-ratio arms of the relevance ablation (G = 32, 50k context)
rq3_abl_nat <- list.files("results/runs_ablation", pattern = "\\.parquet$",
                          full.names = TRUE) |>
  lapply(nanoparquet::read_parquet) |>
  bind_rows() |>
  filter(ratio_spec %in% "natural", G == 32) |>
  group_by(dataset, arm) |>
  summarise(pr = mean(pr_auc), .groups = "drop")
rq3_abl_of <- \(ds, a) sprintf("%.3f",
  rq3_abl_nat$pr[rq3_abl_nat$dataset == ds & rq3_abl_nat$arm == a])

## design-space sweep: PR-AUC and legit score by forced group count
rq3_rapds <- list.files("results/runs_rapds", pattern = "\\.parquet$",
                        full.names = TRUE) |>
  lapply(nanoparquet::read_parquet) |>
  bind_rows() |>
  filter(ds_version == 1, n_estimators == 4)

rq3_g <- rq3_rapds |>
  group_by(dataset, model, granularity) |>
  summarise(`PR-AUC` = mean(pr_auc),
            `Mean legitimate score` = mean(legit_score), .groups = "drop") |>
  tidyr::pivot_longer(c(`PR-AUC`, `Mean legitimate score`),
                      names_to = "measure", values_to = "value") |>
  mutate(dataset = factor(dataset, c("eu_cc", "banksim", "paysim"),
                          c("EU CC", "BankSim", "PaySim")),
         model = factor(model, rq3_md_lv, rq3_md_lab),
         measure = factor(measure, c("PR-AUC", "Mean legitimate score")))
         
rq3_refs <- v4 |>
  filter(condition %in% c("C1", "C2"), model %in% rq3_md_lv,
         seed %in% c(7, 42, 123),
         dataset %in% c("eu_cc", "banksim", "paysim")) |>
  group_by(dataset, condition) |>
  summarise(pr = mean(pr_auc), .groups = "drop") |>
  mutate(dataset = factor(dataset, c("eu_cc", "banksim", "paysim"),
                          c("EU CC", "BankSim", "PaySim")),
         measure = factor("PR-AUC", c("PR-AUC", "Mean legitimate score")))
```

The primary finding to the third research question is negative. Filling the context by kNN retrieval at the natural fraud ratio (C3) scores worse than the random context (C1) on four of the five datasets, sometimes with a catastrophic fall on the performance. The mean PR-AUC delta is for TabPFN v3  `r rq3_d_of("paysim", "tabpfn_3")` on paysim and `r rq3_d_of("eu_cc", "tabpfn_3")` on eu_cc, and respectively for TabICL v2 `r rq3_d_of("paysim", "tabiclv2")` and `r rq3_d_of("eu_cc", "tabiclv2")`. The banksim drop is mild for TabPFN v3 `r rq3_d_of("banksim", "tabpfn_3")` and for TabICL 2 `r rq3_d_of("banksim", "tabiclv2")`, and the fifar contrast of `r rq3_d_of("fifar", "tabpfn_3")` and `r rq3_d_of("fifar", "tabiclv2")` rests on the single pre-defined split. All contrasts in this section use only the same seeded runs experiment C3 ran. Experiment C1 carried ten seeds for these two models shown on @fig-rq1-c1-prauc, so unmatched means would mix the seed sets. On baf the outcome is a tie with deltas of `r rq3_d_of("baf", "tabpfn_3", 3)` and `r rq3_d_of("baf", "tabiclv2", 3)`.

The per-seed spread is itself a finding. On eu_cc the worst seed falls to a PR-AUC of `r rq3_c3_of("eu_cc", min)` while the best stays close to C1 at `r rq3_c3_of("eu_cc", max)`. On paysim every seed collapses, from `r rq3_c3_of("paysim", max)` down to `r rq3_c3_of("paysim", min)`.

A drop under C3 does not by itself isolate relevance, because the realized fraud share of a retrieved context drifts with the local neighbourhood and the silhouette criterion chooses the grouping (@sec-methods-conditions). The ratio-matched relevance ablation of @sec-app-relevance-ablation therefore carries the verdict on RQ3 it holds the grouping fixed at G=32 and enforces the fraud ratio, so only the selection mechanism varies. 

The grouping the silhouette criterion picks under C3 is coarse: on average `r rq3_G_of("banksim")` groups on banksim, `r rq3_G_of("paysim")` on paysim and `r rq3_G_of("fifar")` on fifar, `r rq3_G_of("eu_cc")` on eu_cc and `r rq3_G_of("baf")` on baf. Cluster quality is weak outside banksim, with mean silhouette scores of `r rq3_sil_of("banksim")` on banksim against `r rq3_sil_of("paysim")` on paysim, `r rq3_sil_of("eu_cc")` on eu_cc, `r rq3_sil_of("fifar")` on fifar and `r rq3_sil_of("baf")` on baf. Coarseness alone does not explain the loss, however. We can see in @fig-rq3-granularity that forcing finer groupings in the design-space sweep recovers part of it but asymptotes below C2 on all six dataset and model pairs; @sec-results-rq4 quantifies the sweep's outcomes.


```{r}
#| label: fig-rq3-granularity
#| fig-cap: "The Relevance sweep under specified group counts, means pooled over metrics, ratios and seeds. Top row has PR-AUC recovering with finer G on every dataset but staying below the matched-seed C2 mean (dashed) on all six dataset and model pairs; on BankSim the C1 and C2 references coincide. Bottom row has the mean predicted score on legitimate (non-fraud) rows falling with finer G everywhere, so retrieved-context legit inflation vanishes at fine G on EU CC and BankSim while PaySim remains collapsed, ruling that mechanism out as the PaySim explanation."
#| fig-width: 6.5
#| fig-height: 4.6
#| warning: false

ggplot(rq3_g, aes(granularity, value, colour = model, shape = model)) +
  geom_hline(data = rq3_refs, aes(yintercept = pr, linetype = condition),
             colour = "#52514e", linewidth = 0.4) +
  geom_line(aes(group = model), linewidth = 0.6) +
  geom_point(size = 1.8) +
  facet_grid(measure ~ dataset, scales = "free_y", switch = "y") +
  scale_colour_manual(values = rq3_cols, name = NULL) +
  scale_shape_manual(values = rq3_shapes, name = NULL) +
  scale_linetype_manual(values = c(C1 = "dotted", C2 = "dashed"), name = NULL) +
  scale_x_continuous(trans = "log2", breaks = c(8, 16, 32, 64)) +
  labs(x = "Forced group count G (log scale)", y = NULL) +
  theme_minimal(base_size = 10) +
  theme(legend.position = "bottom", panel.grid.minor = element_blank(),
        strip.placement = "outside")
```

The ablation and the design-space sweep cover eu_cc, banksim and paysim only, so for baf and fifar the evidence is limited to the contrasts of the main experiments. The per-dataset answer to RQ3 is thus negative but graded: retrieval-relevance is strongly harmful on paysim, mildly harmful on banksim, neutral on eu_cc once the ratio and the grouping are controlled, a tie at the floor on baf, and directional only on fifar. This is a regime and domain transfer gap rather than a general refutation of retrieval. Frozen kNN retrieval helps on balanced general tabular data [@thomas2024localpfn_retrieval_finetuning], while under fraud-grade imbalance our result confirms the strong-random-baseline prior [@feuer2023scaling_tabpfn] and that naïve distance-based context construction methods can affect performance [@pham2026aware].

## Research question 4: efficacy of the full RAP-pipeline {#sec-results-rq4}

```{r}
#| label: rq4-setup
#| include: false

## seed-matched condition pivot; needs fraud_ratio, which results-setup drops
rq4_sel <- c("dataset", "model", "condition", "seed", "pipeline_version",
             "n_estimators", "pr_auc", "recall_at_1fpr", "recall_at_5fpr",
             "fraud_ratio")
rq4_all <- list.files("results/runs", pattern = "\\.parquet$",
                      full.names = TRUE) |>
  lapply(\(f) { d <- nanoparquet::read_parquet(f)
                d[, intersect(rq4_sel, names(d)), drop = FALSE] }) |>
  bind_rows() |>
  filter(pipeline_version == 4, n_estimators == 4,
         dataset %in% formal_datasets, model %in% rq3_md_lv)
rq4_cmp <- rq4_all |>
  semi_join(rq4_all |> filter(condition == "C4") |> distinct(dataset, seed),
            by = c("dataset", "seed"))
rq4_base <- rq4_cmp |> filter(condition %in% c("C1", "C2", "C3")) |>
  group_by(dataset, condition, model) |>
  summarise(pr = mean(pr_auc), .groups = "drop") |>
  group_by(dataset, condition) |>
  summarise(pr = max(pr), .groups = "drop")
rq4_c4b <- rq4_cmp |> filter(condition == "C4") |>
  group_by(dataset, model, fraud_ratio) |>
  summarise(pr = mean(pr_auc), .groups = "drop") |>
  group_by(dataset) |>
  summarise(pr = max(pr), .groups = "drop") |>
  mutate(condition = "C4best")
rq4_piv <- bind_rows(rq4_base, rq4_c4b) |>
  tidyr::pivot_wider(names_from = condition, values_from = pr) |>
  mutate(top = pmax(C1, C2, C3, C4best),
         winner = case_when(top == C2 ~ "C2", top == C1 ~ "C1",
                            top == C3 ~ "C3", TRUE ~ "C4"),
         best = if_else(winner %in% c("C1", "C2") & abs(C2 - C1) < 0.005,
                        "C2 ≈ C1", winner),
         dataset = factor(dataset, formal_datasets,
                          c("EU CC", "BankSim", "PaySim", "FiFAR", "BAF"))) |>
  arrange(dataset)
rq4_fmt <- \(x) ifelse(x < 0.095, sprintf("%.3f", x), sprintf("%.2f", x))
rq4_piv_of <- \(ds, col) rq4_fmt(rq4_piv[[col]][rq4_piv$dataset == ds])
rq4_def_of <- \(ds) rq4_fmt(rq4_piv$C2[rq4_piv$dataset == ds] -
                            rq4_piv$C4best[rq4_piv$dataset == ds])

## design-space sweep aggregates (reuses rq3_rapds from rq3-setup)
rq4_gc <- rq3_rapds |> group_by(dataset, granularity) |>
  summarise(pr = mean(pr_auc), .groups = "drop")
rq4_g_of <- \(ds, g) sprintf("%.2f",
  rq4_gc$pr[rq4_gc$dataset == ds & rq4_gc$granularity == g])
rq4_met <- rq3_rapds |> group_by(dataset, metric) |>
  summarise(pr = mean(pr_auc), .groups = "drop")
rq4_met_of <- \(ds, m) sprintf("%.2f",
  rq4_met$pr[rq4_met$dataset == ds & rq4_met$metric == m])
rq4_best <- rq3_rapds |>
  group_by(dataset, model, metric, granularity, fraud_ratio) |>
  summarise(pr = mean(pr_auc), .groups = "drop") |>
  group_by(dataset) |> slice_max(pr, n = 1, with_ties = FALSE) |> ungroup()
rq4_best_of <- \(ds) sprintf("%.3f", rq4_best$pr[rq4_best$dataset == ds])

## per-model C2 means, for the same-model reading of the best swept cell
rq4_c2m <- rq4_cmp |> filter(condition == "C2") |>
  group_by(dataset, model) |>
  summarise(pr = mean(pr_auc), .groups = "drop")
rq4_c2m_of <- \(ds, m) sprintf("%.3f",
  rq4_c2m$pr[rq4_c2m$dataset == ds & rq4_c2m$model == m])

## per-model C4 ranking fragility
rq4_frag <- rq4_cmp |> filter(condition == "C4") |>
  group_by(model) |>
  summarise(n = n(), zero1 = sum(recall_at_1fpr == 0), .groups = "drop")
rq4_frag_of <- \(m, col) rq4_frag[[col]][rq4_frag$model == m]

## guard: the C4-below-C2 conclusion must also hold on both recall metrics,
## as claimed in prose; a future data change that breaks it fails the render
for (rq4_m in c("recall_at_1fpr")) {
  rq4_c2r <- rq4_cmp |> filter(condition == "C2") |>
    group_by(dataset, model) |>
    summarise(v = mean(.data[[rq4_m]]), .groups = "drop") |>
    group_by(dataset) |> summarise(v = max(v), .groups = "drop")
  rq4_c4r <- rq4_cmp |> filter(condition == "C4") |>
    group_by(dataset, model, fraud_ratio) |>
    summarise(v = mean(.data[[rq4_m]]), .groups = "drop") |>
    group_by(dataset) |> summarise(v = max(v), .groups = "drop")
  stopifnot(all(rq4_c4r$v < rq4_c2r$v))
}

## heatmap cells: means over seeds and both metrics
rq4_hm <- rq3_rapds |>
  group_by(dataset, model, granularity, fraud_ratio) |>
  summarise(pr = mean(pr_auc), .groups = "drop") |>
  mutate(dataset = factor(dataset, c("eu_cc", "banksim", "paysim"),
                          c("EU CC", "BankSim", "PaySim")),
         model = factor(model, rq3_md_lv, rq3_md_lab),
         G = factor(granularity, c(8, 16, 32, 64)),
         ratio = factor(fraud_ratio, c(0.05, 0.1, 0.2, 0.3, 0.5)))
```

Research question 4 asks whether the full RAP pipeline (C4), which combines kNN retrieval with an enforced in-context fraud ratio, improves on what rebalancing alone could achieve. In short,it does not. Even with its fraud ratio tuned per dataset, C4 stays below the stratified random context C2 on every dataset, with a decline in performance that ranges from `r rq4_def_of("EU CC")` PR-AUC on eu_cc to `r rq4_def_of("PaySim")` on paysim. The full changes in changes performance can be seen in @tbl-rq4-pivot.

```{r}
#| label: tbl-rq4-pivot
#| tbl-cap: "Seed-matched PR-AUC by context-construction condition. Each value is the better of TabPFN v3 and TabICLv2, averaged over the seeds C4 ran; the C4 column takes the best swept fraud ratio per dataset. This selection favours C4, which still loses to C2 everywhere. FiFAR rests on its single split."

rq4_piv |>
  mutate(across(c(C1, C3, C4best), \(x) sprintf("%.3f", x)),
         C2 = sprintf("**%.3f**", C2)) |>
  select(dataset, C1, C2, C3, C4best, best) |>
  gt() |>
  fmt_markdown(columns = C2) |>
  cols_label(dataset = "Dataset", C4best = "C4 (best ratio)",
             best = "Best strategy") |>
  cols_align("left", columns = dataset) |>
  cols_align("right", columns = c(C1, C2, C3, C4best)) |>
  cols_align("center", columns = best)
```

The deployed pipeline is partly held back by its grouping. As previously @sec-results-rq3 showed, the automatic silhouette choice picks too few groups, so retrieval concentrates around a nearly global centroid. The design-space sweep of @sec-methods-grid supplies the missing information by forcing the group count and sweeping the fraud ratio and the distance metric on the three separable datasets. @fig-rq4-sweep-heatmap maps the full space. We can see that both models perform similarly, so the baheviour is not model specific.

```{r}
#| label: fig-rq4-sweep-heatmap
#| fig-cap: "Seed-mean PR-AUC over the RAP design space. Forced group count $G$ against the enforced fraud ratio, models as rows and datasets as columns. Cells pool three seeds and both distance metrics on a shared 0 to 1 fill scale. No plotted cell reaches its dataset's C2 mean; the one metric-resolved cell that comes within the tie band of C2 (TabICLv2 with cosine on EU CC) is discussed in the text."
#| fig-width: 6.5
#| fig-height: 4.4
#| warning: false

ggplot(rq4_hm, aes(G, ratio, fill = pr)) +
  geom_tile(colour = "#fcfcfb", linewidth = 0.6) +
  geom_text(aes(label = sub("^0\\.", ".", sprintf("%.2f", pr)),
                colour = pr > 0.55), size = 2.4, show.legend = FALSE) +
  facet_grid(model ~ dataset) +
  scale_fill_gradient(low = "#cde2fb", high = "#0d366b", name = "PR-AUC",
                      limits = c(0, 1)) +
  scale_colour_manual(values = c(`TRUE` = "#fcfcfb", `FALSE` = "#0b0b0b")) +
  labs(x = "Forced group count G", y = "Enforced fraud ratio") +
  theme_minimal(base_size = 10) +
  theme(legend.position = "bottom", panel.grid = element_blank(),
        legend.key.width = unit(1.4, "cm"),
        legend.key.height = unit(0.3, "cm"))
```

@sec-results-rq3 already established that finer groupings recover part of the loss but asymptote below C2, the sweep quantifies the outcome. On banksim the pooled mean rises from `r rq4_g_of("banksim", 8)` at $G = 8$ to `r rq4_g_of("banksim", 64)` at $G = 64$ against a C2 mean of `r rq4_piv_of("BankSim", "C2")`. On eu_cc the recovery peaks at `r rq4_g_of("eu_cc", 32)` at $G = 32$ against `r rq4_piv_of("EU CC", "C2")`. On paysim the collapse persists at every grouping, from `r rq4_g_of("paysim", 8)` to `r rq4_g_of("paysim", 64)` against `r rq4_piv_of("PaySim", "C2")`. Across the sweep, cosine distance beats euclidean on banksim (`r rq4_met_of("banksim", "cosine")` against `r rq4_met_of("banksim", "euclidean")`) and eu_cc (`r rq4_met_of("eu_cc", "cosine")` against `r rq4_met_of("eu_cc", "euclidean")`) and sits within the tie band on paysim, with euclidean nominally ahead (`r rq4_met_of("paysim", "euclidean")` against `r rq4_met_of("paysim", "cosine")`). This settles the open metric choice in favour of cosine. Eventhough, the better metric still loses to C2.

Even the best single configuration approaches C2 on one dataset only. The strongest cell is TabICLv2 with cosine retrieval in every case: $G = 64$ at a ratio of 0.05 reaches `r rq4_best_of("eu_cc")` on eu_cc, within the tie band of its C2 mean of `r rq4_piv_of("EU CC", "C2")`, and `r rq4_best_of("banksim")` on banksim against `r rq4_piv_of("BankSim", "C2")`, while paysim peaks at `r rq4_best_of("paysim")` ($G = 16$, ratio 0.50) far below `r rq4_piv_of("PaySim", "C2")`. The winning configurations share a pattern: fine groupings, cosine distance and, on the two near-separable datasets, the lowest swept ratio.

Two further checks sharpen the verdict. The better-of-two pivot hides a model split: under C4, TabPFN v3 scores a Recall at 1% FPR of exactly zero on `r rq4_frag_of("tabpfn_3", "zero1")` of its `r rq4_frag_of("tabpfn_3", "n")` runs, against none of TabICLv2's `r rq4_frag_of("tabiclv2", "n")`, so part of the collapse behind the C4 cells is a model-specific ranking fragility rather than a property of retrieval as such. The conclusion is also not metric-specific: the best C4 stays below C2 on every dataset on Recall at 1% FPR and Recall at 5% FPR as well. The answer to RQ4 is therefore negative. Ratio tuning on top of retrieval does not rescue RAP, and no swept configuration justifies retrieval over plain stratified balancing in the tested datasets.

## Comparison with state-of-the-art fraud detectors {#sec-results-sota}

```{r}
#| label: sota-setup
#| include: false

## grid slice: C2 for the two strongest TFMs, all available seeds
sota_sel <- c("dataset", "model", "condition", "seed", "pipeline_version",
              "n_estimators", "pr_auc", "recall_at_1fpr", "recall_at_5fpr",
              "roc_auc")
sota_grid <- list.files("results/runs", pattern = "\\.parquet$",
                        full.names = TRUE) |>
  lapply(\(f) { d <- nanoparquet::read_parquet(f)
                d[, intersect(sota_sel, names(d)), drop = FALSE] }) |>
  bind_rows() |>
  filter(pipeline_version == 4, n_estimators == 4 | is.na(n_estimators),
         dataset %in% formal_datasets)
sota_lab <- c(tabpfn_3 = "TabPFN v3", tabiclv2 = "TabICLv2")
sota_c2 <- sota_grid |>
  filter(condition == "C2", model %in% names(sota_lab)) |>
  group_by(dataset, model) |>
  summarise(pr = mean(pr_auc), pr_sd = sd(pr_auc),
            r1 = mean(recall_at_1fpr), r5 = mean(recall_at_5fpr),
            roc = mean(roc_auc), .groups = "drop")
sota_best_pr <- sota_c2 |> group_by(dataset) |>
  slice_max(pr, n = 1, with_ties = FALSE) |> ungroup()
sota_best_r5 <- sota_c2 |> group_by(dataset) |>
  slice_max(r5, n = 1, with_ties = FALSE) |> ungroup()
fmt3 <- \(x) sprintf("%.3f", x)
sota_pr_of   <- \(ds) fmt3(sota_best_pr$pr[sota_best_pr$dataset == ds])
sota_prm_of  <- \(ds) sota_lab[sota_best_pr$model[sota_best_pr$dataset == ds]]
sota_prsd_of <- \(ds) fmt3(sota_best_pr$pr_sd[sota_best_pr$dataset == ds])
sota_r5_of   <- \(ds) fmt3(sota_best_r5$r5[sota_best_r5$dataset == ds])
sota_r5m_of  <- \(ds) sota_lab[sota_best_r5$model[sota_best_r5$dataset == ds]]
sota_c2_of   <- \(ds, m) fmt3(sota_c2$pr[sota_c2$dataset == ds & sota_c2$model == m])
sota_m_of    <- \(ds, m, col) fmt3(
  sota_c2[[col]][sota_c2$dataset == ds & sota_c2$model == m])

## eu_cc cross-validation study (results/runs_sota_cv; n_estimators = 12,
## outside the grid, never mixed into grid pivots)
sota_cv <- list.files("results/runs_sota_cv", pattern = "\\.parquet$",
                      full.names = TRUE) |>
  lapply(nanoparquet::read_parquet) |> bind_rows()
sota_cv_sum <- sota_cv |> group_by(model, arm) |>
  summarise(folds = n(), n_splits = first(n_splits),
            pr = mean(pr_auc), pr_sd = sd(pr_auc),
            dup = mean(dup_factor), .groups = "drop")
sota_cv_of <- \(m, a, col) {
  r <- sota_cv_sum |> filter(model == m, arm == a); r[[col]] }
sota_ctx <- scales::comma(round(mean(sota_cv$context_size)))

## pooled metrics for complete natural arms: every fraud scored exactly once
## across the folds; tie-grouped average precision, stratified bootstrap
sota_ap <- function(y, s) {
  o <- order(s, decreasing = TRUE); y <- y[o]; s <- s[o]
  g <- cumsum(!duplicated(s)); tp <- cumsum(y); nn <- seq_along(y)
  last <- !duplicated(g, fromLast = TRUE)
  prec <- tp[last] / nn[last]; rec <- tp[last] / sum(y)
  sum(diff(c(0, rec)) * prec)
}
sota_cv_complete <- sota_cv_sum |>
  filter(arm == "natural", folds == n_splits) |> pull(model)
## the bootstrap takes ~1 min, so it is cached to disk keyed by the run ids,
## seed and rep count; new folds change the ids and force a recompute
sota_pool <- lapply(sota_cv_complete, \(m) {
  ids <- sort(sota_cv |> filter(model == m, arm == "natural") |> pull(run_id))
  cache_dir <- "results/runs_sota_cv/pooled_cache"
  dir.create(cache_dir, showWarnings = FALSE)
  cache_f <- file.path(cache_dir, paste0(
    m, "_", substr(rlang::hash(list(ids, 1000, 42, "v2")), 1, 12), ".rds"))
  if (file.exists(cache_f)) return(readRDS(cache_f))
  p <- bind_rows(lapply(
    file.path("results/runs_sota_cv/preds", paste0(ids, ".parquet")),
    nanoparquet::read_parquet))
  y <- p$y_true; s <- p$prob
  ip <- which(y == 1); ineg <- which(y == 0)
  set.seed(42)
  bs <- replicate(1000, {
    i <- c(sample(ip, replace = TRUE), sample(ineg, replace = TRUE))
    sota_ap(y[i], s[i]) })
  r_at <- \(t) {                       # conservative step, as in metrics.py
    o <- order(s, decreasing = TRUE); yy <- y[o]
    fp <- cumsum(1 - yy); tp <- cumsum(yy)
    max(c(0, (tp / sum(yy))[fp / sum(1 - yy) <= t])) }
  res <- tibble(model = m, n_fraud = sum(y), ap = sota_ap(y, s),
                lo = quantile(bs, 0.025), hi = quantile(bs, 0.975),
                r1 = r_at(0.01), r5 = r_at(0.05),
                roc = (mean(rank(s)[y == 1]) - (sum(y) + 1) / 2) / sum(1 - y))
  saveRDS(res, cache_f)
  res
}) |> bind_rows()
sota_pool_of <- \(m, col) fmt3(sota_pool[[col]][sota_pool$model == m])
sota_nfraud <- sota_pool$n_fraud[1]

## external comparators (verified figures; sources cited in prose).
## andradearenas2025 reports single-split point values (RF 0.871, XGBoost
## 0.867); its bootstrap CIs cover accuracy/recall/F1 only, not AUPRC, so the
## external figure columns carry no whiskers (no cross-source borrowing)
ext_eucc <- 0.871
ext_eucc_xgb <- 0.867

## guards: the verdicts claimed in prose must hold on the current data
stopifnot(sota_best_pr$pr[sota_best_pr$dataset == "banksim"] > 0.772)
stopifnot(sota_pool$lo[sota_pool$model == "tabpfn_3"] < ext_eucc,
          ext_eucc < sota_pool$hi[sota_pool$model == "tabpfn_3"],
          sota_pool$lo[sota_pool$model == "tabiclv2"] < ext_eucc,
          ext_eucc < sota_pool$hi[sota_pool$model == "tabiclv2"])
```

This final subchapter benchmarks the strongest TFM results of this study against the strongest published results on the same datasets. The comparison is restricted to methodologically sound external work. We accept an external figure as a benchmark only if it reports an imbalance-aware metric, keeps the test set at its natural class ratio and applies any resampling strictly inside the training partition. Much of the published literature on these datasets fails one or more of these filters. These publications usually then produce report near-perfect figures, AUPRC or ROC-AUC of 0.99 and above. This unrealistic performance typically is a result from rebalancing the test set or from applying SMOTE [@chawla2002smote] before splitting. @tbl-sota-comparison collects, per dataset, our collected metrics for both frontier TFM models separately in their best-case configuration, next to the best verified external figures on the metrics their sources report.

| Dataset | Model | PR-AUC | R@1%FPR | R@5%FPR | ROC-AUC |
|:--|:--|--:|--:|--:|--:|
| eu_cc | TabPFN v3 | `r sota_pool_of("tabpfn_3", "ap")` | `r sota_pool_of("tabpfn_3", "r1")` | `r sota_pool_of("tabpfn_3", "r5")` | `r sota_pool_of("tabpfn_3", "roc")` |
| | TabICLv2 | `r sota_pool_of("tabiclv2", "ap")` | `r sota_pool_of("tabiclv2", "r1")` | `r sota_pool_of("tabiclv2", "r5")` | `r sota_pool_of("tabiclv2", "roc")` |
| | External [@andradearenas2025smote_credit_card_comparison] | 0.871 | | | 0.978 |
| banksim | TabPFN v3 | `r sota_m_of("banksim", "tabpfn_3", "pr")` | `r sota_m_of("banksim", "tabpfn_3", "r1")` | `r sota_m_of("banksim", "tabpfn_3", "r5")` | `r sota_m_of("banksim", "tabpfn_3", "roc")` |
| | TabICLv2 | `r sota_m_of("banksim", "tabiclv2", "pr")` | `r sota_m_of("banksim", "tabiclv2", "r1")` | `r sota_m_of("banksim", "tabiclv2", "r5")` | `r sota_m_of("banksim", "tabiclv2", "roc")` |
| | External [@hirnschall2025logsig_bayesian_gan_fraud] | 0.772 | | | |
| paysim | TabPFN v3 | `r sota_m_of("paysim", "tabpfn_3", "pr")` | `r sota_m_of("paysim", "tabpfn_3", "r1")` | `r sota_m_of("paysim", "tabpfn_3", "r5")` | `r sota_m_of("paysim", "tabpfn_3", "roc")` |
| | TabICLv2 | `r sota_m_of("paysim", "tabiclv2", "pr")` | `r sota_m_of("paysim", "tabiclv2", "r1")` | `r sota_m_of("paysim", "tabiclv2", "r5")` | `r sota_m_of("paysim", "tabiclv2", "roc")` |
| | External [@visbeek2023deep_symbolic_classification_fraud] | | | | |
| fifar | TabPFN v3 | `r sota_m_of("fifar", "tabpfn_3", "pr")` | `r sota_m_of("fifar", "tabpfn_3", "r1")` | `r sota_m_of("fifar", "tabpfn_3", "r5")` | `r sota_m_of("fifar", "tabpfn_3", "roc")` |
| | TabICLv2 | `r sota_m_of("fifar", "tabiclv2", "pr")` | `r sota_m_of("fifar", "tabiclv2", "r1")` | `r sota_m_of("fifar", "tabiclv2", "r5")` | `r sota_m_of("fifar", "tabiclv2", "roc")` |
| | External [@alves2023fifar_learning_to_defer] | | | 0.579 | |
| baf | TabPFN v3 | `r sota_m_of("baf", "tabpfn_3", "pr")` | `r sota_m_of("baf", "tabpfn_3", "r1")` | `r sota_m_of("baf", "tabpfn_3", "r5")` | `r sota_m_of("baf", "tabpfn_3", "roc")` |
| | TabICLv2 | `r sota_m_of("baf", "tabiclv2", "pr")` | `r sota_m_of("baf", "tabiclv2", "r1")` | `r sota_m_of("baf", "tabiclv2", "r5")` | `r sota_m_of("baf", "tabiclv2", "roc")` |
| | External [@jesus2022baf_arxiv] | | | 0.45 to 0.50 | |

: All collected ranking metrics for the two strongest TFMs in the best case scenarios from this study next to the best verified external figures. The eu_cc rows use the full-context protocol. A pooled predictions of the five-fold cross-validation described in the text. The values of the other datasets are seed-means of the stratified balanced context (C2) runs, whose 48,000 to 50,000-row context limit makes them conservative and should be rather considered as a lower bound for performance. Empty cells in the external rows are metrics the source does not report. For example, recall at a fixed FPR appears in no verified source on eu_cc, banksim or paysim. Furthermore, paysim's only sound external figure is a thresholded F1 of 0.82, not comparable to any column. The banksim source also reports its own deviation of $\pm$ 0.010. {#tbl-sota-comparison}

The EU credit card datasets comparison is the closest to like-for-like, since both report PR-AUC over the full ULB dataset at its natural class prevalence. On this dataset we therefore report the full-context protocol. A stratified five-fold cross-validation over all 284,807 rows, with the entire training fold of about `r sota_ctx` rows as the context and an ensemble size of 12.  Pooling the fold predictions, so that each of the `r sota_nfraud` frauds is scored exactly once, TabPFN v3 at the natural in-context fraud ratio reaches a PR-AUC of `r sota_pool_of("tabpfn_3", "ap")`, with a stratified bootstrap 95 % confidence interval of `r sota_pool_of("tabpfn_3", "lo")` to `r sota_pool_of("tabpfn_3", "hi")`. TabICLv2 reaches a pooled PR-AUC of `r sota_pool_of("tabiclv2", "ap")`, with an interval of `r sota_pool_of("tabiclv2", "lo")` to `r sota_pool_of("tabiclv2", "hi")`, nominally ahead of TabPFN v3. The SOTA results from literature are 0.871 for a tuned random forest and 0.867 for XGBoost [@andradearenas2025smote_credit_card_comparison]. Another benchmark without resampling corroborates that level with a kNN PR-AUC of 0.867 $\pm$ 0.033 under repeated cross-validation [@brownlee2020imbalanced_creditcard_tutorial]. Our point values are nominally the highest of these figures, but both confidence intervals comfortably contain the external points. Results can be interpreted that the zero-shot TFMs match the verified state of the art on this dataset, not that they beat it. A second arm that tops the context fraud share up to 2 % by duplication averages `r fmt3(sota_cv_of("tabpfn_3", "ratio_0.02", "pr"))` for TabPFN v3 and `r fmt3(sota_cv_of("tabiclv2", "ratio_0.02", "pr"))` for TabICLv2 over the folds, no better than the natural ratio for either model, consistent with the duplication penalty of @sec-results-rq2.

<!-- TODO: Add internal link to c1 below -->

```{r}
#| label: fig-sota-eucc
#| fig-cap: "PR-AUC on eu_cc, naive against optimised, per model. Solid columns are the naive result: a random C1 context for the TFMs and our full-data fit for the GBDT baselines. The light extension raises each column to the model's best optimised result. For the frontier TFMs this is the full training context and the stratified balanced context C2 for the rest. For the tree models, the optimised values are the best verified external figures, a tuned random forest and tuned XGBoost on a single split [@andradearenas2025smote_credit_card_comparison]. The random forest has no naive counterpart in this study, so its column is a single optimised layer. Whiskers extend one standard deviation above the mean. The external values carry no whiskers as their source reports no confidence intervals for AUPRC."
#| fig-width: 6.5
#| fig-height: 3.8
#| warning: false

sota_lab_map <- c(rf_external = "Random forest", xgboost = "XGBoost",
                  catboost = "CatBoost", tabpfn_v2 = "TabPFN v2",
                  tabpfn_25 = "TabPFN 2.5", tabpfn_26 = "TabPFN 2.6",
                  tabpfn_3 = "TabPFN v3", tabiclv2 = "TabICLv2")
sota_naive <- sota_grid |> filter(dataset == "eu_cc", condition == "C1") |>
  group_by(model) |>
  summarise(value = mean(pr_auc), sd = sd(pr_auc), .groups = "drop") |>
  mutate(layer = "Naive")
## optimised = CV natural arm when its five folds exist, C2 grid mean otherwise;
## the CV rows take over automatically once the remaining folds are run
sota_opt_cv <- sota_cv |>
  filter(arm == "natural", model %in% sota_cv_complete) |>
  group_by(model) |>
  summarise(value = mean(pr_auc), sd = sd(pr_auc), .groups = "drop")
sota_opt <- sota_grid |>
  filter(dataset == "eu_cc", condition == "C2") |>
  group_by(model) |>
  summarise(value = mean(pr_auc), sd = sd(pr_auc), .groups = "drop") |>
  anti_join(sota_opt_cv, by = "model") |>
  bind_rows(sota_opt_cv) |>
  mutate(layer = "Optimised")
sota_ext <- tibble(model = c("rf_external", "xgboost"),
                   value = c(ext_eucc, ext_eucc_xgb),
                   sd = NA_real_, layer = "Optimised")
sota_fig <- bind_rows(sota_ext, sota_opt, sota_naive) |>  # optimised drawn behind
  mutate(family = if_else(model %in% c(gbdt_models, "rf_external"),
                          "Tree", "TFM"),
         layer = factor(layer, c("Naive", "Optimised")))
sota_ord <- sota_fig |> group_by(model) |>
  summarise(top = max(value), .groups = "drop") |>
  arrange(desc(top)) |> pull(model)
sota_fig <- sota_fig |>
  mutate(model_lab = factor(model, sota_ord, sota_lab_map[sota_ord]))
sota_whisk <- sota_fig |> group_by(model_lab) |>
  slice_max(value, n = 1, with_ties = FALSE)

ggplot(sota_fig, aes(model_lab, value, fill = model_lab, alpha = layer)) +
  geom_col_pattern(aes(pattern = family), position = "identity", width = 0.7,
                   colour = "grey25", linewidth = 0.15,
                   pattern_colour = NA, pattern_fill = "white",
                   pattern_angle = 45, pattern_density = 0.25,
                   pattern_spacing = 0.02, pattern_key_scale_factor = 0.6) +
  geom_errorbar(data = sota_whisk, aes(ymin = value, ymax = value + sd),
                width = 0.3, linewidth = 0.35, colour = "grey25",
                alpha = 1, na.rm = TRUE) +
  scale_fill_manual(values = c(
    "Random forest" = "grey55", "XGBoost" = "#2a78d6",
    "CatBoost" = "#008300", "TabPFN v2" = "#e87ba4",
    "TabPFN 2.5" = "#eda100", "TabPFN 2.6" = "#1baf7a",
    "TabPFN v3" = "#eb6834", "TabICLv2" = "#4a3aa7")) +
  scale_alpha_manual(values = c(Naive = 1, Optimised = 0.4), name = NULL) +
  scale_pattern_manual(values = c(Tree = "stripe", TFM = "none")) +
  scale_y_continuous(limits = c(0, 1.02), expand = expansion(0)) +
  labs(x = NULL, y = "PR-AUC") +
  guides(fill = "none", pattern = "none",
         alpha = guide_legend(override.aes = list(fill = "grey35",
                                                  pattern = "none"))) +
  theme_minimal(base_size = 10) +
  theme(legend.position = "bottom", panel.grid.minor = element_blank(),
        panel.grid.major.x = element_blank(), axis.ticks.x = element_blank())
```

@fig-sota-eucc visualises the effects on optimising the context. Balancing alone lifts most TFMs close to the external tree benchmark, and the final full training context run lifts both frontier TFMs above them, TabICLv2 now narrowly ahead of all other models. On the tree side the naïve columns are our fixed-default full-data fits (@sec-methods-models) and the optimised values come from the external source, so the gap between the solid XGBoost column and its light extension is roughly the price of forgoing per-dataset hyperparameter tuning.

On banksim our `r sota_pr_of("banksim")` (`r sota_prm_of("banksim")`) clearly exceeds the best verified external PR-AUC of 0.772, reported for random forest and feed-forward baselines in a semi-supervised benchmark [@hirnschall2025logsig_bayesian_gan_fraud]. In this comparison there is two caveats that cut in opposite directions, so this cannot be considered as one-to-one compariso. The external figure is a deliberately label-scarce operating point, with 5 % of training labels revealed, so a fully supervised external ensemble could plausibly score higher. On the other hand, our split is grouped by customer as was elaborated in @sec-methods-splits while the external split is class-stratified only. So our protocol is a stricter one and customer leakage cannot be ruled out on their results. No published BankSim benchmark with a customer-disjoint split was found.

The remaining two datasets are compared on recall at a 5 % false-positive rate. On fifar we reach `r sota_r5_of("fifar")` against the 0.579 of the LightGBM alert model shipped with the dataset [@alves2023fifar_learning_to_defer]. The external figure is a validation-fold value with the operating threshold chosen on the same fold, and it is the only standalone classification result published on fifar, so we read the comparison as at par. On baf we reach `r sota_r5_of("baf")` against the 0.45 to 0.50 read from the scatter of tuned LightGBM configurations in the original benchmark [@jesus2022baf_arxiv], corroborated by an independent reproduction at 0.450 [@nasif2026snn_hyperheuristic_fraud] and a peer-reviewed runner-up at 0.471 [@perdigao2024snn_joint_activity_fraud]. Our value is nominally above that band, but the published figures use the canonical temporal split while our grid uses random stratified splits, so the temporal ablation of @sec-app-second is the honest comparable cell.

Across the five datasets the pattern is uniform. A zero-shot TFM with a stratified balanced context, frozen weights, no per-dataset tuning and a deliberately reduced ensemble (@sec-methods-models) is above the sound external state of the art on banksim, at par on eu_cc and fifar, nominally above on baf under a different split protocol, and without a sound comparator on paysim. None of the surveyed sources on eu_cc, banksim or paysim reports recall at a fixed false-positive rate, so the recall columns of our grid double as a reference benchmark for that gap. Three caveats bound the claim. The random-split protocol keeps time as a feature and may inflate the ranking metrics of the separable datasets (@sec-methods-splits). The per-dataset choice of the better TFM is mildly in our favour. For increased clarity @tbl-sota-comparison reports both models separately.
