---
bibliography: ../thesis-refs.bib
---

# CC_2025 as a cautionary tale {#sec-app-cc .appendix}

```{r}
#| label: cc-app-setup
#| include: false

# cc_2025 is shipped in FULL in data/sample/ (small enough), so this is the
# complete dataset at true class balance, not a fraud-overrepresented sample
# (see data/sample/README.md). Inside thesis.qmd these packages are already
# attached by 4_data.qmd; reloading them is a no-op and keeps the appendix
# renderable standalone.
suppressPackageStartupMessages({
  library(ggplot2)
  library(dplyr)
  library(scales)
  library(patchwork)
  library(gt)
  library(glue)
  library(nanoparquet)
})

cc_2025 <- nanoparquet::read_parquet("data/sample/cc_2025.parquet")
names(cc_2025) <- tolower(names(cc_2025))

cc_rate <- mean(cc_2025$fraud_flag)
cc_country <- summarise(cc_2025, rate = mean(fraud_flag), .by = country)
cc_country_p <- chisq.test(table(cc_2025$country, cc_2025$fraud_flag))$p.value
cc_type_p <- chisq.test(table(cc_2025$transaction_type, cc_2025$fraud_flag))$p.value

pct <- function(x, acc = 0.01) percent(x, accuracy = acc, suffix = " %")

# ATM / Online / POS series colours (CVD-validated triple)
col_txn <- c(ATM = "#2563EB", Online = "#D97706", POS = "#7C3AED")

cap_cc_country <- glue(
  "Fraud rate by country. The per-country rates span ",
  "{pct(min(cc_country$rate))} to {pct(max(cc_country$rate))} around the ",
  "global rate of {pct(cc_rate, 0.1)} (dashed line); the differences are not ",
  "statistically significant (chi-squared p = {round(cc_country_p, 2)})."
)
cap_cc_type <- glue(
  "Fraud rate by transaction type. The three rates agree with the global ",
  "rate of {pct(cc_rate, 0.1)} (dashed line) to two decimal places ",
  "(chi-squared p = {round(cc_type_p, 2)})."
)
```

- the labels seem to be statistically independent on every single feature.

```{r}
#| label: fig-app-cc-country
#| fig-cap: !expr cap_cc_country
#| fig-width: 6.2
#| fig-height: 2.6
#| warning: false
#| message: false

cc_country |>
  ggplot(aes(reorder(country, rate), rate)) +
  geom_col(fill = "grey65", width = 0.7) +
  geom_hline(yintercept = cc_rate, linetype = "dashed",
             colour = "grey30", linewidth = 0.4) +
  scale_y_continuous(labels = label_percent(suffix = " %")) +
  labs(x = NULL, y = "Fraud rate") +
  theme_minimal(base_size = 10) +
  theme(panel.grid.minor = element_blank(),
        panel.grid.major.x = element_blank())
```

```{r}
#| label: fig-app-cc-day
#| fig-cap: "Hourly structure of cc_2025. Transaction volume is uniform over the day (top), and the fraud rate of every transaction type fluctuates only within sampling noise around the global rate (dashed line, bottom). Real payment data shows a strong diurnal cycle in both."
#| fig-width: 6.2
#| fig-height: 4.6
#| warning: false
#| message: false

by_hour <- cc_2025 |>
  summarise(n = n(), rate = mean(fraud_flag),
            .by = c(hour_of_day, transaction_type))

p_volume <- by_hour |>
  summarise(n = sum(n), .by = hour_of_day) |>
  ggplot(aes(hour_of_day, n)) +
  geom_col(width = 0.8, fill = "grey65") +
  scale_y_continuous(labels = label_number(scale = 1e-3, suffix = "k")) +
  labs(x = NULL, y = "Transactions") +
  theme_minimal(base_size = 10) +
  theme(panel.grid.minor = element_blank())

p_rate <- by_hour |>
  ggplot(aes(hour_of_day, rate, colour = transaction_type)) +
  geom_hline(yintercept = cc_rate, linetype = "dashed",
             colour = "grey30", linewidth = 0.4) +
  geom_line(linewidth = 0.6) +
  scale_colour_manual(values = col_txn, name = NULL) +
  scale_y_continuous(labels = label_percent(suffix = " %"),
                     limits = c(0, NA)) +
  labs(x = "Hour of day", y = "Fraud rate") +
  theme_minimal(base_size = 10) +
  theme(panel.grid.minor = element_blank())

(p_volume / p_rate) + plot_layout(guides = "collect") &
  theme(legend.position = "bottom")
```

```{r}
#| label: fig-app-cc-type
#| fig-cap: !expr cap_cc_type
#| fig-width: 6.2
#| fig-height: 2.4
#| warning: false
#| message: false

cc_2025 |>
  summarise(rate = mean(fraud_flag), .by = transaction_type) |>
  ggplot(aes(reorder(transaction_type, rate), rate)) +
  geom_col(fill = "grey65", width = 0.6) +
  geom_hline(yintercept = cc_rate, linetype = "dashed",
             colour = "grey30", linewidth = 0.4) +
  scale_y_continuous(labels = label_percent(suffix = " %")) +
  labs(x = NULL, y = "Fraud rate") +
  theme_minimal(base_size = 10) +
  theme(panel.grid.minor = element_blank(),
        panel.grid.major.x = element_blank())
```


```{r}
#| label: tbl-app-cc-device
#| tbl-cap: "Chip and PIN usage by device type (exact full-data counts; cc_2025 is shipped whole in data/sample/). The share of every chip/PIN combination is identical across device types, including chip use recorded for about a quarter of Web transactions, and the fraud rate is flat across all twelve cells."

cc_2025 |>
  summarise(n = n(), rate = mean(fraud_flag),
            .by = c(device_type, is_chip, is_pin_used)) |>
  mutate(share = n / sum(n), .by = device_type) |>
  arrange(device_type, is_chip, is_pin_used) |>
  mutate(across(c(is_chip, is_pin_used),
                \(x) ifelse(x == 1, "Yes", "No"))) |>
  select(device_type, is_chip, is_pin_used, n, share, rate) |>
  gt(groupname_col = "device_type") |>
  cols_label(
    is_chip = "Chip",
    is_pin_used = "PIN",
    n = "Rows",
    share = "Share",
    rate = "Fraud rate"
  ) |>
  fmt_number(columns = n, decimals = 0, use_seps = TRUE) |>
  fmt_percent(columns = c(share, rate), decimals = 1)
```

All in all seems that there has been  

# Time split ablations {#sec-app-second .appendix}

# The fraud-ratio and context-size sweep {#sec-app-c2-sweep .appendix}

Condition C2 fixes a single balancing target, but the choice of operating point matters as much as the mechanism itself. This sweep maps the stratified-random mechanism over the full plane of fraud ratio and context size, so the ratio effect can be characterised as a curve rather than read off a single point. It also gives the C4 ratio sweep (@sec-methods-conditions) a like-for-like counterpart at every ratio: one condition with retrieval, one without, both under the same with-replacement fraud policy.

The core sweep crosses nine fraud ratios (2, 4, 5, 6, 8, 10, 20, 30 and 40 percent) with five requested context sizes (10,000, 20,000, 30,000, 40,000 and 50,000 rows). An extension run on 2026-07-20 adds requested sizes of 100,000 and 200,000 rows at the four ratios shared with the core sweep (2, 5, 10 and 20 percent), through the same driver and the same fraud policy. The effective context size is capped at the smaller of the training-set size and the model's checkpoint limit, and both the requested and the effective size are recorded; only EU CC approaches this cap, its training split of about 228,000 rows nearly saturating the 200,000-row request. Two TFMs (TabICLv2 and TabPFN v3) are run on three datasets (EU CC, BankSim and PaySim), each at three seeds, with the same ensemble size (n_estimators = 4) as the main grid, for 810 core runs and 144 extension runs. Each cell fits a single global context and one model refit; there is no per-group retrieval or refitting here, unlike C3/C4. At the extended sizes TabPFN v3 stays in-spec: its checkpoint accepts 1,000,000 rows, so the 50,000-row budget of the main grid is a design choice rather than the model cap. TabICLv2 has no hard cap, but 48,000 rows is its pre-training upper bound, so its 100,000- and 200,000-row cells measure extrapolation beyond the pre-training range and must be read as such. These three datasets span the fraud-scarcity range of the formal grid: EU CC holds only about 394 training fraud rows, while BankSim and PaySim hold thousands of them, a difference that matters for the next point.

```{r}
#| label: c2-app-setup
#| include: false

c2_grid <- list.files("results/runs_c2grid", pattern = "\\.parquet$",
                      full.names = TRUE) |>
  lapply(nanoparquet::read_parquet) |>
  bind_rows()

# one point per sweep cell: mean over the 2 models x 3 seeds, over the full
# 10k-200k plane (core sweep + the 2026-07-20 extension). The directory also
# holds the n_estimators sweep (incl. baf/fifar rows); the n_estimators == 4
# and dataset filters keep only this sweep's cells (the shared n = 4 runs at
# 10 %/50k are the same run files, not duplicates).
c2_cells <- c2_grid |>
  filter(n_estimators == 4,
         dataset %in% c("banksim", "eu_cc", "paysim")) |>
  summarise(pr_auc = mean(pr_auc), dup_factor = mean(dup_factor),
            .by = c(dataset, model, fraud_ratio, context_size_req)) |>
  mutate(ctx = factor(context_size_req,
                      labels = c("10k", "20k", "30k", "40k", "50k",
                                 "100k", "200k")))

# context-size ramp shared by both sweep figures
col_ctx <- c("10k" = "#BFDBFE", "20k" = "#8FC0FC", "30k" = "#60A5FA",
             "40k" = "#4283F2", "50k" = "#2563EB", "100k" = "#214EBA",
             "200k" = "#1E3A8A")

# per-model size means at the four shared ratios, for the extension prose
c2_ext <- c2_grid |>
  filter(n_estimators == 4, dataset %in% c("banksim", "eu_cc", "paysim"),
         fraud_ratio %in% c(0.02, 0.05, 0.10, 0.20)) |>
  summarise(pr_auc = mean(pr_auc), .by = c(dataset, model, context_size_req))

c2_bs_v3 <- filter(c2_ext, dataset == "banksim", model == "tabpfn_3")
c2_bs_50 <- c2_bs_v3$pr_auc[c2_bs_v3$context_size_req == 50000]
c2_bs_200 <- c2_bs_v3$pr_auc[c2_bs_v3$context_size_req == 200000]
c2_ps_icl <- filter(c2_ext, dataset == "paysim", model == "tabiclv2")
c2_ps_gain <- max(c2_ps_icl$pr_auc[c2_ps_icl$context_size_req > 50000]) -
  c2_ps_icl$pr_auc[c2_ps_icl$context_size_req == 50000]
c2_eu_rng <- c2_ext |>
  filter(dataset == "eu_cc", context_size_req >= 30000) |>
  summarise(pr_auc = mean(pr_auc), .by = context_size_req) |>
  pull(pr_auc) |>
  range()

cap_c2_dup <- glue(
  "Cell-mean PR-AUC against the duplication factor (log axis; each point is ",
  "one ratio and context cell of @fig-app-c2-ratio, and lines connect the ",
  "cells of one context size across the swept ratios). Cells at a ",
  "duplication factor of one realise their ratio without duplication. Lines ",
  "that cross a factor of roughly two decline towards their high-duplication ",
  "end. Only EU CC reaches heavy duplication, up to about ",
  "{round(max(c2_cells$dup_factor))}-fold; its larger contexts sit above the ",
  "smaller ones at a matched duplication factor but do not escape the ",
  "decline. Holding context size fixed removes the confound between context ",
  "size and duplication: within a line, moving right means raising the ratio."
)
```

@fig-app-c2-ratio shows the seed- and model-mean PR-AUC over the swept plane.

```{r}
#| label: fig-app-c2-ratio
#| fig-cap: "Mean PR-AUC over the fraud ratio and context size plane, averaged over both models and all seeds (log-scaled ratio axis; the vertical scale differs per panel). The 100,000- and 200,000-row contexts run only at the four ratios shared with the core sweep, so their lines are shorter. Larger contexts help mainly at low ratios and where fraud is scarcest (EU CC). Ratios beyond about 10 % degrade EU CC and BankSim, while PaySim plateaus at mid ratios; the steepest declines are the extended contexts on BankSim."
#| fig-width: 6.2
#| fig-height: 6.2
#| warning: false
#| message: false

c2_cells |>
  ggplot(aes(fraud_ratio, pr_auc, colour = ctx, group = ctx)) +
  geom_line(linewidth = 0.6) +
  geom_point(size = 1.2) +
  scale_colour_manual(values = col_ctx, name = "Context size") +
  scale_x_log10(breaks = c(0.02, 0.05, 0.10, 0.20, 0.40),
                labels = label_percent(suffix = " %")) +
  labs(x = "Requested in-context fraud ratio", y = "PR-AUC") +
  theme_minimal(base_size = 10) +
  theme(legend.position = "bottom", panel.grid.minor = element_blank())+
  facet_grid(rows=vars(model), cols =vars(dataset), scales = "free_y")
```

The context's fraud quota is filled by sampling with replacement once unique training fraud rows run out, the same policy used by the C4 sampler. This has two consequences. First, the requested ratio is always realised, so a high-ratio cell measures the intended ratio rather than a silently smaller one produced by exhausting the available fraud rows. Second, duplication becomes an observable and recorded quantity, the ratio of fraud rows in the context to unique fraud rows among them. On EU CC the arithmetic forces heavy duplication at high ratios: a 40 percent quota of a 50,000-row context is 20,000 fraud rows drawn against roughly 394 unique frauds, close to 50-fold duplication, and the extended contexts push this to about 25-fold already at 100,000 rows and 10 percent and to about `r round(max(c2_cells$dup_factor))`-fold at 200,000 rows and 20 percent. BankSim and PaySim, by contrast, stay near a duplication factor of one through the low and mid ratios, reach only roughly three-to-3.5-fold at the core sweep's highest ratios (on BankSim's ~5,760 training frauds and PaySim's ~6,570), and roughly six-to-sevenfold at the largest extended contexts, more than an order of magnitude milder than EU CC. The sweep therefore doubles as a duplication study conducted entirely in a random context. Any effect of duplication alone appears here, free of retrieval, which gives the later analysis of C4 behaviour a control that separates duplication effects from retrieval effects. @fig-app-c2-dup plots the same cells against their duplication factor. <!-- TODO: crossref the results-chapter sweep section once it exists -->

```{r}
#| label: fig-app-c2-dup
#| fig-cap: !expr cap_c2_dup
#| fig-width: 6.2
#| fig-height: 3.6
#| warning: false
#| message: false

c2_cells |>
  ggplot(aes(dup_factor, pr_auc, colour = ctx, group = ctx)) +
  geom_line(linewidth = 0.6) +
  geom_point(size = 1.2) +
  facet_grid(rows=vars(model), cols =vars(dataset), scales = "free_x") +
  scale_colour_manual(values = col_ctx, name = "Context size") +
  #scale_x_log10(breaks = c(1, 3, 10, 30, 100)) +
  labs(x = "Duplication factor (context fraud rows per unique fraud row)",
       y = "PR-AUC") +
  theme_minimal(base_size = 10) +
  theme(legend.position = "bottom", panel.grid.minor = element_blank())
```

The extension answers the budget question directly: relaxing the 50,000-row design budget of the main grid buys nothing. On EU CC the model-mean PR-AUC at the four shared ratios lands between `r sprintf("%.3f", c2_eu_rng[1])` and `r sprintf("%.3f", c2_eu_rng[2])` at every size from 30,000 rows upward. On PaySim the best extended context improves on the 50,000-row one by `r sprintf("%.3f", c2_ps_gain)` at most (TabICLv2). BankSim declines monotonically once the context grows past 50,000 rows, on both models; TabPFN v3 falls from `r sprintf("%.3f", c2_bs_50)` at 50,000 rows to `r sprintf("%.3f", c2_bs_200)` at 200,000 rows (means over the four shared ratios).

@fig-app-c2-dup also collapses the interaction between context size and ratio into one pattern. Cells that grow duplication-free, the stacks at a duplication factor of one and in practice the 2 percent cells on BankSim and PaySim, keep improving out to 200,000 rows. Lines that cross a factor of roughly two instead decline towards their high-duplication end, and on EU CC the larger contexts sit above the smaller ones at a matched duplication factor without escaping that decline. This is consistent with a simple reading: at a fixed ratio, a larger context multiplies duplicated fraud rows, so high-ratio cells scale the pathology with size while low-ratio cells scale information. <!-- TODO: per-model breakdown of the extension; the duplication reading above is a candidate mechanism, not yet pinned down -->

# Ensemble size sweep {#sec-app-nest .appendix}

- n_estimators swept over {1, 2, 4, 8, 10, 12} at the fixed plateau recipe (10 percent fraud ratio, 50,000-row context), both TFMs, all five formal datasets (13 dataset-seed units, 156 cells; the n = 4 cells are shared with the sweep of @sec-app-c2-sweep). Run 2026-07-20.
- Within each unit the context rows are identical across n_estimators values (the sampling RNG does not include it), so the sweep isolates the effect of ensembling on a fixed context.
- Curves are flat from n = 2 upward and saturate by n = 8. The main grid's settled n_estimators = 4 gives up at most about 0.006 PR-AUC against n = 8 (EU CC / TabPFN v3), typically at most 0.002; n = 10 and n = 12 add nothing over n = 8.
- n = 1 is visibly worse where ensembling matters: PaySim / TabICLv2 at 0.966 against 0.980 at saturation, EU CC / TabPFN v3 at 0.813 against about 0.827.
- Predict time scales roughly linearly with n_estimators, so the settled n = 4 halves inference cost against the package default of 8 at negligible accuracy cost.
- TODO: quantify the accuracy-cost trade-off with the recorded fit_s / predict_s columns.

```{r}
#| label: nest-setup
#| include: false

nest_cells <- c2_grid |>
  filter(fraud_ratio == 0.10, context_size_req == 50000) |>
  summarise(pr_auc = mean(pr_auc), .by = c(dataset, model, n_estimators)) |>
  mutate(model = c(tabiclv2 = "TabICLv2", tabpfn_3 = "TabPFN v3")[model])
```

```{r}
#| label: fig-app-nest
#| fig-cap: "Seed-mean PR-AUC against the number of ensemble passes at the fixed plateau recipe (10 % fraud ratio, 50k context; the vertical scale differs per panel). Curves saturate by n = 8, and the main grid's n_estimators = 4 sits within about 0.006 of the saturated value everywhere."
#| fig-width: 6.2
#| fig-height: 4.2
#| warning: false
#| message: false

nest_cells |>
  ggplot(aes(n_estimators, pr_auc, colour = model)) +
  geom_line(linewidth = 0.6) +
  geom_point(size = 1.2) +
  facet_wrap(~dataset, nrow = 2, scales = "free_y") +
  scale_colour_manual(values = c("TabPFN v3" = "#eb6834",
                                 "TabICLv2" = "#4a3aa7"), name = NULL) +
  scale_x_continuous(breaks = c(1, 2, 4, 8, 10, 12)) +
  labs(x = "Ensemble passes (n_estimators)", y = "PR-AUC") +
  theme_minimal(base_size = 10) +
  theme(legend.position = "bottom", panel.grid.minor = element_blank())
```

# Ratio-matched relevance ablation {#sec-app-relevance-ablation .appendix}

Neither grid contrast isolates relevance cleanly. The step from C1 to C3 varies the selection mechanism but lets the realised fraud ratio drift with the local neighbourhood of each cluster, so a ratio effect and a relevance effect are entangled. The step from C3 to C4 changes ratio enforcement and the retrieval mode at the same time, so it is not a single-factor comparison either. The ablation described here removes both confounds by crossing the two selection mechanisms independently, at a matched ratio, context size and granularity.

The design is a 2x2 over selection mechanisms: legitimate rows drawn either at random or by kNN, crossed with fraud rows drawn either at random or by kNN. This yields four arms: RR (random legitimate rows, random fraud rows), KR (kNN legitimate rows, random fraud rows), RK (random legitimate rows, kNN fraud rows) and KK (kNN both). All four arms run per group at a fixed granularity of G = 32 k-means clusters of the scored test set. Granularity is fixed rather than silhouette-selected here, so grouping is held constant across arms and cannot itself explain a difference between them. Retrieval uses the cosine metric, the requested context size is 50,000 rows, and the fraud ratio is swept over 5 percent and 10 percent, filled with the same with-replacement top-up policy as C2 and C4. The ablation runs on TabICLv2 and TabPFN v3, on EU CC, BankSim and PaySim, at three seeds. RR reproduces C2's stratified-random selection, operated at a set ratio rather than the 50/50 target and with the same with-replacement policy as the sweep, under per-group refitting, and KK is the class-conditional retrieval core of C4; each single-letter flip from one of these arms isolates one selector. Because RR runs at the same G = 32 as the other three arms, comparing it against the global-context sweep of @sec-app-c2-sweep additionally isolates the effect of per-group granularity itself, separately from the effect of any selector.

The ablation is additionally run at a natural ratio level, defined per dataset as the training-split fraud prevalence, for the RR and KK arms only. This supplies the clean form of the C1-to-C3 contrast: both arms see the same natural class ratio and the same grouping, so only the selection mechanism differs between them. At the natural level the fraud quota sits below the unique fraud count by construction, so no duplication occurs. <!-- TODO: inline R for the realised natural ratios per dataset -->

Each contrast in this design answers a distinct question. The ratio effect is addressed in @sec-methods-conditions. Granularity is addressed by RR against the global sweep of @sec-app-c2-sweep. Legitimate-selector relevance is addressed by RK against KK and by RR against KR. Fraud-selector relevance is addressed by RR against RK and by KR against KK. The selection mechanism at natural prevalence, free of ratio enforcement, is addressed by RR against KK at the natural level.
