---
bibliography: ../thesis-refs.bib
---

# Degeneracy of the cc_2025 dataset {#sec-app-cc .appendix}

```{r}
#| label: cc-app-setup
#| include: false

# cc_2025 is shipped in FULL in data/sample/ (small enough), so this is the
# complete dataset at true class balance, not a fraud-overrepresented sample
# (see data/sample/README.md). Inside thesis.qmd these packages are already
# attached by 4_data.qmd; reloading them is a no-op and keeps the appendix
# renderable standalone.
suppressPackageStartupMessages({
  library(ggplot2)
  library(dplyr)
  library(scales)
  library(patchwork)
  library(gt)
  library(glue)
  library(nanoparquet)
})

cc_2025 <- nanoparquet::read_parquet("data/sample/cc_2025.parquet")
names(cc_2025) <- tolower(names(cc_2025))

cc_rate <- mean(cc_2025$fraud_flag)
cc_country <- summarise(cc_2025, rate = mean(fraud_flag), .by = country)
cc_country_p <- chisq.test(table(cc_2025$country, cc_2025$fraud_flag))$p.value
cc_type_p <- chisq.test(table(cc_2025$transaction_type, cc_2025$fraud_flag))$p.value
cc_amt <- tapply(cc_2025$amount, cc_2025$fraud_flag, mean)

# per-customer country-hopping: the sample ships the identifier columns, so
# customers are reconstructable even though the models never see the IDs.
# cc_ctry_exp = expected distinct countries if each transaction's country is
# an independent uniform draw over the 8 countries, at each customer's own
# transaction count
cc_cust <- summarise(cc_2025, tx = n(), n_ctry = n_distinct(country),
                     .by = customer_id)
cc_ctry_exp <- mean(8 * (1 - (7 / 8)^cc_cust$tx))

# feature-wise independence tests against the label: chi-squared for the
# categorical and binary features, two-sample Kolmogorov-Smirnov for the
# continuous ones (the rounded amounts produce ties, hence the suppressed
# approximate-p warning; immaterial at n = 500,000)
cc_cat <- c("merchant_category", "card_type", "transaction_type", "country",
            "device_type", "is_international", "is_chip", "is_pin_used")
cc_num <- c("amount", "distance_from_home", "hour_of_day")
cc_p_all <- c(
  sapply(cc_cat, \(v) chisq.test(table(cc_2025[[v]], cc_2025$fraud_flag))$p.value),
  sapply(cc_num, \(v) suppressWarnings(
    ks.test(cc_2025[[v]][cc_2025$fraud_flag == 1],
            cc_2025[[v]][cc_2025$fraud_flag == 0])$p.value))
)
cc_p_min <- min(cc_p_all)

# the archived cc_2025 grid runs (results/runs_cc2025_degenerate; kept as
# documentation of the degeneracy after the dataset was dropped)
cc_runs <- list.files("results/runs_cc2025_degenerate", pattern = "\\.parquet$",
                      full.names = TRUE) |>
  lapply(\(f) nanoparquet::read_parquet(f)[, c("model", "condition",
                                               "pr_auc", "roc_auc")]) |>
  bind_rows()
cc_pr_rng <- range(cc_runs$pr_auc)
cc_roc_rng <- range(cc_runs$roc_auc)

pct <- function(x, acc = 0.01) percent(x, accuracy = acc, suffix = " %")

# ATM / Online / POS series colours (CVD-validated triple)
col_txn <- c(ATM = "#2563EB", Online = "#D97706", POS = "#7C3AED")

cap_cc_country <- glue(
  "Fraud rate by country. The per-country rates span ",
  "{pct(min(cc_country$rate))} to {pct(max(cc_country$rate))} around the ",
  "global rate of {pct(cc_rate, 0.1)} (dashed line); the differences are not ",
  "statistically significant (chi-squared p = {round(cc_country_p, 2)})."
)
cap_cc_type <- glue(
  "Fraud rate by transaction type. The three rates agree with the global ",
  "rate of {pct(cc_rate, 0.1)} (dashed line) to two decimal places ",
  "(chi-squared p = {round(cc_type_p, 2)})."
)
```

Why should the CC_2025 dataset be rejected from the evalutions [@rajak2025credit_card_fraud_2025]? First of all, there is no statistical difference between fraudulend and legtimate transactions. Both have the same average amount (`r sprintf("%.1f", cc_amt["1"])` against `r sprintf("%.1f", cc_amt["0"])`) and occurs at the same distance from the cardholder's home. No country is riskier than another. @fig-app-cc-country shows the per-country fraud rates spanning only `r pct(min(cc_country$rate))` to `r pct(max(cc_country$rate))` around the global `r pct(cc_rate, 0.1)`. @fig-app-cc-type shows ATM, online and point-of-sale transactions to be fraudulent at rates identical to two decimal places, and in the hourly profile of @fig-app-cc-day, fraud at four in the morning is no more common than fraud at noon. Nothing in the data distinguishes the two classes. The fraud flag behaves as if it were attached to each transaction by an independent coin flip of probability `r pct(cc_rate, 0.1)`.

Formal tests confirm this impression feature by feature. For each of the eight categorical and binary features, a chi-squared test of the feature-label contingency table finds no association; for the three continuous features (amount, distance from home and hour of day), a two-sample Kolmogorov-Smirnov test finds no difference between the fraud and non-fraud distributions. The smallest p-value over all eleven tests is `r sprintf("%.2f", cc_p_min)` (device type), unremarkable among eleven tests under the null. At half a million rows these tests would detect minuscule real effects, so the absence of signal is not a power problem. The labels are statistically independent of every feature.

```{r}
#| label: fig-app-cc-country
#| fig-cap: !expr cap_cc_country
#| fig-width: 6.2
#| fig-height: 2.6
#| warning: false
#| message: false

cc_country |>
  ggplot(aes(reorder(country, rate), rate)) +
  geom_col(fill = "grey65", width = 0.7) +
  geom_hline(yintercept = cc_rate, linetype = "dashed",
             colour = "grey30", linewidth = 0.4) +
  scale_y_continuous(labels = label_percent(suffix = " %")) +
  labs(x = NULL, y = "Fraud rate") +
  theme_minimal(base_size = 10) +
  theme(panel.grid.minor = element_blank(),
        panel.grid.major.x = element_blank())
```

The degeneracy extends beyond the labels to the generative process itself. @fig-app-cc-day shows the hourly structure: transaction volume is uniform over the day, and the fraud rate of every transaction type fluctuates only within sampling noise around the global rate. The informative datasets of the formal grid behave nothing like this; @fig-fraud-rate-time in the data chapter shows their fraud shares drifting, spiking and cycling over their native time columns. The uniformity recurs in every marginal: the three transaction types and the three device types each cover almost exactly a third of the rows, a giveaway that the columns were sampled independently from flat distributions rather than generated by any transaction process.

```{r}
#| label: fig-app-cc-day
#| fig-cap: "Hourly structure of cc_2025. Transaction volume is uniform over the day (top), and the fraud rate of every transaction type fluctuates only within sampling noise around the global rate (dashed line, bottom). Real payment data shows a strong diurnal cycle in both."
#| fig-width: 6.2
#| fig-height: 4.6
#| warning: false
#| message: false

by_hour <- cc_2025 |>
  summarise(n = n(), rate = mean(fraud_flag),
            .by = c(hour_of_day, transaction_type))

p_volume <- by_hour |>
  summarise(n = sum(n), .by = hour_of_day) |>
  ggplot(aes(hour_of_day, n)) +
  geom_col(width = 0.8, fill = "grey65") +
  scale_y_continuous(labels = label_number(scale = 1e-3, suffix = "k")) +
  labs(x = NULL, y = "Transactions") +
  theme_minimal(base_size = 10) +
  theme(panel.grid.minor = element_blank())

p_rate <- by_hour |>
  ggplot(aes(hour_of_day, rate, colour = transaction_type)) +
  geom_hline(yintercept = cc_rate, linetype = "dashed",
             colour = "grey30", linewidth = 0.4) +
  geom_line(linewidth = 0.6) +
  scale_colour_manual(values = col_txn, name = NULL) +
  scale_y_continuous(labels = label_percent(suffix = " %"),
                     limits = c(0, NA)) +
  labs(x = "Hour of day", y = "Fraud rate") +
  theme_minimal(base_size = 10) +
  theme(panel.grid.minor = element_blank())

(p_volume / p_rate) + plot_layout(guides = "collect") &
  theme(legend.position = "bottom")
```

```{r}
#| label: fig-app-cc-type
#| fig-cap: !expr cap_cc_type
#| fig-width: 6.2
#| fig-height: 2.4
#| warning: false
#| message: false

cc_2025 |>
  summarise(rate = mean(fraud_flag), .by = transaction_type) |>
  ggplot(aes(reorder(transaction_type, rate), rate)) +
  geom_col(fill = "grey65", width = 0.6) +
  geom_hline(yintercept = cc_rate, linetype = "dashed",
             colour = "grey30", linewidth = 0.4) +
  scale_y_continuous(labels = label_percent(suffix = " %")) +
  labs(x = NULL, y = "Fraud rate") +
  theme_minimal(base_size = 10) +
  theme(panel.grid.minor = element_blank(),
        panel.grid.major.x = element_blank())
```


Independent column-wise sampling also produces internal contradictions that no real payment system could emit. The customer identifier makes this vivid. The `r label_comma()(nrow(cc_cust))` customers place `r sprintf("%.1f", mean(cc_cust$tx))` transactions each on average, and those few transactions visit `r sprintf("%.2f", mean(cc_cust$n_ctry))` of the eight countries on average, matching to two decimals the `r sprintf("%.2f", cc_ctry_exp)` expected if every transaction's country were an independent uniform draw. The typical cardholder thus hops between four countries within a handful of purchases, yet only `r pct(mean(cc_2025$is_international), 1)` of transactions carry the international flag, because that column too is sampled independently of everything else. @tbl-app-cc-device shows the same incoherence in the payment mechanics, crossing device type with the chip and PIN indicators. Every chip/PIN combination holds an identical share within each device type, including chip use recorded for about 30 % of Web transactions, although a physical chip read cannot occur in a browser. The fraud rate is flat across all twelve cells. A dataset whose columns contradict each other in this way cannot encode a joint distribution for a model to learn, whatever its label quality.

```{r}
#| label: tbl-app-cc-device
#| tbl-cap: "Chip and PIN usage by device type (exact full-data counts; cc_2025 is shipped whole in data/sample/). The share of every chip/PIN combination is identical across device types, including chip use recorded for about 30 % of Web transactions, and the fraud rate is flat across all twelve cells."

cc_2025 |>
  summarise(n = n(), rate = mean(fraud_flag),
            .by = c(device_type, is_chip, is_pin_used)) |>
  mutate(share = n / sum(n), .by = device_type) |>
  arrange(device_type, is_chip, is_pin_used) |>
  mutate(across(c(is_chip, is_pin_used),
                \(x) ifelse(x == 1, "Yes", "No"))) |>
  select(device_type, is_chip, is_pin_used, n, share, rate) |>
  gt(groupname_col = "device_type") |>
  cols_label(
    is_chip = "Chip",
    is_pin_used = "PIN",
    n = "Rows",
    share = "Share",
    rate = "Fraud rate"
  ) |>
  fmt_number(columns = n, decimals = 0, use_seps = TRUE) |>
  fmt_percent(columns = c(share, rate), decimals = 1)
```

Taken together, the evidence supports a simple account of how the dataset was made: each column, the fraud flag included, was drawn independently from its own marginal distribution. Under that construction no model can be better than chance. The cautionary tale is aimed at benchmark selection. The dataset is publicly distributed, superficially plausible, and passes every casual inspection: realistic column names, sensible value ranges, a credible 1.5 % fraud rate. Its degeneracy only becomes visible on contact, through independence tests that cost seconds, or through model runs that could cost hours. Any above-chance result reported on this dataset can therefore only reflect an evaluation error, such as target leakage or duplicated rows across the train-test split, and results published on unaudited synthetic benchmarks should be read with that possibility in mind. The pre-flight checks demonstrated here, per-feature association tests against the label and a scan of the marginals for implausible uniformity, are cheap enough to run before any compute is spent, and this study now applies them as a standing admission criterion for benchmark data.

# Time-split ablation {#sec-app-second .appendix}

```{r}
#| label: ts-app-setup
#| include: false

ts_runs <- list.files("results/runs_timesplit", pattern = "\\.parquet$",
                      full.names = TRUE) |>
  lapply(nanoparquet::read_parquet) |>
  bind_rows()

# matched random-split control: the plateau-recipe cells of the sweep in
# @sec-app-c2-sweep (10 % ratio, 50k request, n_estimators = 4, same models
# and seeds). Loaded here independently of c2-app-setup because this section
# renders first.
ts_ctl <- list.files("results/runs_c2grid", pattern = "\\.parquet$",
                     full.names = TRUE) |>
  lapply(nanoparquet::read_parquet) |>
  bind_rows() |>
  filter(dataset %in% unique(ts_runs$dataset), fraud_ratio == 0.10,
         context_size_req == 50000, n_estimators == 4)

ts_disp <- c(tabiclv2 = "TabICLv2", tabpfn_3 = "TabPFN v3")

ts_tbl <- ts_runs |>
  summarise(ts_pr = mean(pr_auc), ts_sd = sd(pr_auc),
            .by = c(dataset, model)) |>
  left_join(ts_ctl |>
              summarise(ctl_pr = mean(pr_auc), ctl_sd = sd(pr_auc),
                        .by = c(dataset, model)),
            by = c("dataset", "model")) |>
  mutate(delta = ts_pr - ctl_pr,
         verdict = case_when(ts_pr > ctl_pr + ctl_sd ~ "above",
                             ts_pr < ctl_pr - ctl_sd ~ "below",
                             .default = "inside"))

ts_v <- function(d, m, col) ts_tbl[[col]][ts_tbl$dataset == d & ts_tbl$model == m]

ts_split <- distinct(ts_runs, dataset, boundary_t, n_fraud_train, prevalence)
ts_b <- function(d) ts_split$boundary_t[ts_split$dataset == d]
ts_frauds <- function(d) ts_split$n_fraud_train[ts_split$dataset == d]
ts_ps_prev <- ts_split$prevalence[ts_split$dataset == "paysim"]

ts_eu_dup <- max(ts_runs$dup_factor[ts_runs$dataset == "eu_cc"])
ts_bs_d <- mean(ts_tbl$delta[ts_tbl$dataset == "banksim"])
ts_icl_rng <- range(ts_runs$pr_auc[ts_runs$dataset == "paysim" &
                                     ts_runs$model == "tabiclv2"])
ts_r5_floor <- min(c(
  ts_runs$recall_at_5fpr[ts_runs$dataset != "eu_cc"],
  ts_ctl$recall_at_5fpr[ts_ctl$dataset != "eu_cc"]))
ts_r5_eu <- mean(ts_runs$recall_at_5fpr[ts_runs$dataset == "eu_cc"])
ctl_r5_eu <- mean(ts_ctl$recall_at_5fpr[ts_ctl$dataset == "eu_cc"])
```

The experiments in this thesis evaluate eu_cc, banksim and paysim under random splits and deliberately keeps their native time columns as features, delivered in  @sec-methods-splits. This poses a clear temporal-leakage threat. The temporal regularities of @fig-fraud-rate-time are available to every model to 'memorise'. So the grid's figures might not survive a deployment-realistic protocol that trains on the past and scores the future. This ablation bounds the threat by forcing exactly such an experiment desing. Each of the three datasets is re-split at a time-value boundary, train being the first roughly 80 percent of rows by the native time column and the test being the rest of the data. Placing the boundary on a time value rather than a row rank guarantees that no timestamp straddles the split. The realised boundaries fall at second `r label_comma()(ts_b("eu_cc"))` of eu_cc's 172,792-second window, at step `r ts_b("banksim")` of banksim's 180 simulated days, and at step `r ts_b("paysim")` of paysim's 743 hourly steps, the last far below the 80 percent mark of the step range because paysim's transaction volume is front-loaded. baf dataset needs no forced split of its own as fifar is a temporally split variant of baf evaluated under its published protocol [@jesus2022baf_arxiv], so the grid already contains baf's temporal counterpart.

Both frontier TFMs (TabPFN v3 and TabICLv2) rerun on the best performing recipe from @sec-app-c2-sweep unchanged on the temporal training side. A stratified-random context of 50 000 requested rows at a 10 percent fraud ratio, filled with the same with-replacement policy, with four estimators, at seeds 7, 42 and 123. The split itself is deterministic, so the seeds vary only the context subsample and the negative-cap test subsample. The time features are kept, which places every test-side time value outside the training range by construction. The measured change therefore bounds the combined effect of time-feature leakage and genuine temporal shift, which is the quantity the threat statement concerns. The control for this experiment is the same recipe under the random split, namely the 10 percent, 50 000-row cells of @sec-app-c2-sweep at the same seeds. The comparison rule is deliberately plain, if a temporal seed-mean inside the control's mean $\pm$ one standard deviation band counts as no measurable inflation, and a mean outside the band is read as the bound on the threat.

The forced split changes more than the ordering, and three consequences must be kept in mind when reading @tbl-app-timesplit. Firstly,  it replaces banksim's grouped split. The main experiments assigns each customer wholly to one side of the split, as elaborated in @sec-methods-splits. While a temporal boundary necessarily lets customers straddle it, so banksim's temporal cell measures ordering and entity leakage together. Secondly, paysim's late test window is compositionally different by simulator construction, as @fig-fraud-rate-time shows, legitimate traffic thins out towards the end of the simulation, so the temporal test side carries a fraud prevalence of `r pct(ts_ps_prev)` against 0.13 percent overall, and the temporal training side holds only `r label_comma()(ts_frauds("paysim"))` unique frauds against roughly 6,570 under a random split. Thirdly, the duplication regime is matched where it matters, the eu_cc's temporal training side holds `r ts_frauds("eu_cc")` frauds against roughly 394 under the random split, so its 10 percent quota duplicates about `r round(ts_eu_dup)`-fold under both protocols, and duplication cannot explain an eu_cc difference.

@tbl-app-timesplit gives the verdicts. On eu_cc both models land inside the control band, nominally lower by `r sprintf("%.3f", abs(ts_v("eu_cc", "tabpfn_3", "delta")))` to `r sprintf("%.3f", abs(ts_v("eu_cc", "tabiclv2", "delta")))` but well within the control's wide seed spread. Despite a kept time feature and a collection window of only two days, the random split shows no measurable temporal inflation. On banksim both models land above the band, by `r sprintf("%+.3f", ts_bs_d)` on average, meaning the temporal protocol scores higher than the random one. The direction is consistent with the removed customer grouping, since the encoded customer ID can act as a memorisation key across a temporal boundary, which is exactly what the grouped split exists to prevent. Paysim results split by model, TabPFN v3 stays inside the band at `r sprintf("%+.3f", ts_v("paysim", "tabpfn_3", "delta"))`, while TabICLv2 falls `r sprintf("%.3f", abs(ts_v("paysim", "tabiclv2", "delta")))` below it and its seed spread widens from `r sprintf("%.3f", ts_v("paysim", "tabiclv2", "ctl_sd"))` to `r sprintf("%.3f", ts_v("paysim", "tabiclv2", "ts_sd"))`. The split is deterministic and the seeds change only the context sample, so the widened spread means TabICLv2 becomes strongly context-sensitive under this shift while TabPFN v3 does not. 

Summed up, the temporal-leakage threat to the main grid is bounded tightly. No measurable and persistent inflation across datasets and models could be identified.

```{r}
#| label: tbl-app-timesplit
#| tbl-cap: "Temporal-split PR-AUC against the matched random-split control, per dataset and model. Both columns are seed means with one standard deviation over seeds {7, 42, 123} at the shared recipe (10 % fraud ratio, 50,000-row requested context, n_estimators = 4); the control cells are those of the sweep in @sec-app-c2-sweep. The verdict marks whether the temporal mean lies inside the control's mean ± sd band."

ts_tbl |>
  arrange(dataset, model) |>
  mutate(dataset = c(banksim = "BankSim", eu_cc = "EU CC",
                     paysim = "PaySim")[dataset],
         model = ts_disp[model],
         temporal = sprintf("%.3f ± %.3f", ts_pr, ts_sd),
         control = sprintf("%.3f ± %.3f", ctl_pr, ctl_sd),
         delta = sprintf("%+.3f", delta)) |>
  select(dataset, model, temporal, control, delta, verdict) |>
  gt(groupname_col = "dataset") |>
  cols_label(
    model = "Model",
    temporal = "Temporal PR-AUC",
    control = "Random-split PR-AUC",
    delta = "Difference",
    verdict = "Verdict"
  )
```

# The fraud-ratio and context-size sweep {#sec-app-c2-sweep .appendix}

The stratified balancing condition of the main grid fixes a single balancing target, but the choice of operating point matters as much as the mechanism itself. This sweep maps the stratified-random mechanism over the full plane of fraud ratio and context size, so the ratio effect can be characterised as a curve rather than read off a single point. It also gives RAP's ratio sweep, defined in @sec-methods-conditions, a like-for-like counterpart at every ratio: one condition with retrieval, one without, both under the same with-replacement fraud policy.

The core sweep crosses nine fraud ratios (2, 4, 5, 6, 8, 10, 20, 30 and 40 percent) with five requested context sizes (10,000, 20,000, 30,000, 40,000 and 50,000 rows). An extension run on 2026-07-20 adds requested sizes of 100,000 and 200,000 rows at the four ratios shared with the core sweep (2, 5, 10 and 20 percent), through the same driver and the same fraud policy. The effective context size is capped at the smaller of the training-set size and the model's checkpoint limit, and both the requested and the effective size are recorded; only EU CC approaches this cap, its training split of about 228,000 rows nearly saturating the 200,000-row request. Two TFMs (TabICLv2 and TabPFN v3) are run on three datasets (EU CC, BankSim and PaySim), each at three seeds, with the same ensemble size (n_estimators = 4) as the main grid, for 810 core runs and 144 extension runs. Each cell fits a single global context and one model refit; there is no per-group retrieval or refitting here, unlike in the retrieval conditions. At the extended sizes TabPFN v3 stays in-spec: its checkpoint accepts 1,000,000 rows, so the 50,000-row budget of the main grid is a design choice rather than the model cap. TabICLv2 has no hard cap, but 48,000 rows is its pre-training upper bound, so its 100,000- and 200,000-row cells measure extrapolation beyond the pre-training range and must be read as such. These three datasets span the fraud-scarcity range of the formal grid: EU CC holds only about 394 training fraud rows, while BankSim and PaySim hold thousands of them, a difference that matters for the next point.

```{r}
#| label: c2-app-setup
#| include: false

c2_grid <- list.files("results/runs_c2grid", pattern = "\\.parquet$",
                      full.names = TRUE) |>
  lapply(nanoparquet::read_parquet) |>
  bind_rows()

# one point per sweep cell: mean over the 2 models x 3 seeds, over the full
# 10k-200k plane (core sweep + the 2026-07-20 extension). The directory also
# holds the n_estimators sweep (incl. baf/fifar rows); the n_estimators == 4
# and dataset filters keep only this sweep's cells (the shared n = 4 runs at
# 10 %/50k are the same run files, not duplicates).
c2_cells <- c2_grid |>
  filter(n_estimators == 4,
         dataset %in% c("banksim", "eu_cc", "paysim")) |>
  summarise(pr_auc = mean(pr_auc), dup_factor = mean(dup_factor),
            .by = c(dataset, model, fraud_ratio, context_size_req)) |>
  mutate(ctx = factor(context_size_req,
                      labels = c("10k", "20k", "30k", "40k", "50k",
                                 "100k", "200k")))

# context-size ramp shared by both sweep figures
col_ctx <- c("10k" = "#BFDBFE", "20k" = "#8FC0FC", "30k" = "#60A5FA",
             "40k" = "#4283F2", "50k" = "#2563EB", "100k" = "#214EBA",
             "200k" = "#1E3A8A")

# per-model size means at the four shared ratios, for the extension prose
c2_ext <- c2_grid |>
  filter(n_estimators == 4, dataset %in% c("banksim", "eu_cc", "paysim"),
         fraud_ratio %in% c(0.02, 0.05, 0.10, 0.20)) |>
  summarise(pr_auc = mean(pr_auc), .by = c(dataset, model, context_size_req))

c2_bs_v3 <- filter(c2_ext, dataset == "banksim", model == "tabpfn_3")
c2_bs_50 <- c2_bs_v3$pr_auc[c2_bs_v3$context_size_req == 50000]
c2_bs_200 <- c2_bs_v3$pr_auc[c2_bs_v3$context_size_req == 200000]
c2_ps_icl <- filter(c2_ext, dataset == "paysim", model == "tabiclv2")
c2_ps_gain <- max(c2_ps_icl$pr_auc[c2_ps_icl$context_size_req > 50000]) -
  c2_ps_icl$pr_auc[c2_ps_icl$context_size_req == 50000]
c2_eu_rng <- c2_ext |>
  filter(dataset == "eu_cc", context_size_req >= 30000) |>
  summarise(pr_auc = mean(pr_auc), .by = context_size_req) |>
  pull(pr_auc) |>
  range()

cap_c2_dup <- glue(
  "Cell-mean PR-AUC against the duplication factor (log axis; each point is ",
  "one ratio and context cell of @fig-app-c2-ratio, and lines connect the ",
  "cells of one context size across the swept ratios). Cells at a ",
  "duplication factor of one realise their ratio without duplication. Lines ",
  "that cross a factor of roughly two decline towards their high-duplication ",
  "end. Only EU CC reaches heavy duplication, up to about ",
  "{round(max(c2_cells$dup_factor))}-fold; its larger contexts sit above the ",
  "smaller ones at a matched duplication factor but do not escape the ",
  "decline. Holding context size fixed removes the confound between context ",
  "size and duplication: within a line, moving right means raising the ratio."
)
```

@fig-app-c2-ratio shows the seed- and model-mean PR-AUC over the swept plane.

```{r}
#| label: fig-app-c2-ratio
#| fig-cap: "Mean PR-AUC over the fraud ratio and context size plane, averaged over both models and all seeds (log-scaled ratio axis; the vertical scale differs per panel). The 100,000- and 200,000-row contexts run only at the four ratios shared with the core sweep, so their lines are shorter. Larger contexts help mainly at low ratios and where fraud is scarcest (EU CC). Ratios beyond about 10 % degrade EU CC and BankSim, while PaySim plateaus at mid ratios; the steepest declines are the extended contexts on BankSim."
#| fig-width: 6.2
#| fig-height: 6.2
#| warning: false
#| message: false

c2_cells |>
  ggplot(aes(fraud_ratio, pr_auc, colour = ctx, group = ctx)) +
  geom_line(linewidth = 0.6) +
  geom_point(size = 1.2) +
  scale_colour_manual(values = col_ctx, name = "Context size") +
  scale_x_log10(breaks = c(0.02, 0.05, 0.10, 0.20, 0.40),
                labels = label_percent(suffix = " %")) +
  labs(x = "Requested in-context fraud ratio", y = "PR-AUC") +
  theme_minimal(base_size = 10) +
  theme(legend.position = "bottom", panel.grid.minor = element_blank())+
  facet_grid(rows=vars(model), cols =vars(dataset), scales = "free_y")
```

The context's fraud quota is filled by sampling with replacement once unique training fraud rows run out, the same policy used by the RAP sampler. This has two consequences. First, the requested ratio is always realised, so a high-ratio cell measures the intended ratio rather than a silently smaller one produced by exhausting the available fraud rows. Second, duplication becomes an observable and recorded quantity, the ratio of fraud rows in the context to unique fraud rows among them. On EU CC the arithmetic forces heavy duplication at high ratios: a 40 percent quota of a 50,000-row context is 20,000 fraud rows drawn against roughly 394 unique frauds, close to 50-fold duplication, and the extended contexts push this to about 25-fold already at 100,000 rows and 10 percent and to about `r round(max(c2_cells$dup_factor))`-fold at 200,000 rows and 20 percent. BankSim and PaySim, by contrast, stay near a duplication factor of one through the low and mid ratios, reach only roughly three-to-3.5-fold at the core sweep's highest ratios (on BankSim's ~5,760 training frauds and PaySim's ~6,570), and roughly six-to-sevenfold at the largest extended contexts, more than an order of magnitude milder than EU CC. The sweep therefore doubles as a duplication study conducted entirely in a random context, whose ratio curves @sec-results-rq2 reads at the 50,000-row slice. Any effect of duplication alone appears here, free of retrieval, which gives the analysis of RAP behaviour in @sec-results-rq4 a control that separates duplication effects from retrieval effects. @fig-app-c2-dup plots the same cells against their duplication factor.

```{r}
#| label: fig-app-c2-dup
#| fig-cap: !expr cap_c2_dup
#| fig-width: 6.2
#| fig-height: 3.6
#| warning: false
#| message: false

c2_cells |>
  ggplot(aes(dup_factor, pr_auc, colour = ctx, group = ctx)) +
  geom_line(linewidth = 0.6) +
  geom_point(size = 1.2) +
  facet_grid(rows=vars(model), cols =vars(dataset), scales = "free_x") +
  scale_colour_manual(values = col_ctx, name = "Context size") +
  #scale_x_log10(breaks = c(1, 3, 10, 30, 100)) +
  labs(x = "Duplication factor (context fraud rows per unique fraud row)",
       y = "PR-AUC") +
  theme_minimal(base_size = 10) +
  theme(legend.position = "bottom", panel.grid.minor = element_blank())
```

The extension answers the budget question directly: relaxing the 50,000-row design budget of the main grid buys nothing. On EU CC the model-mean PR-AUC at the four shared ratios lands between `r sprintf("%.3f", c2_eu_rng[1])` and `r sprintf("%.3f", c2_eu_rng[2])` at every size from 30,000 rows upward. On PaySim the best extended context improves on the 50,000-row one by `r sprintf("%.3f", c2_ps_gain)` at most (TabICLv2). BankSim declines monotonically once the context grows past 50,000 rows, on both models; TabPFN v3 falls from `r sprintf("%.3f", c2_bs_50)` at 50,000 rows to `r sprintf("%.3f", c2_bs_200)` at 200,000 rows (means over the four shared ratios).

@fig-app-c2-dup also collapses the interaction between context size and ratio into one pattern. Cells that grow duplication-free, the stacks at a duplication factor of one and in practice the 2 percent cells on BankSim and PaySim, keep improving out to 200,000 rows. Lines that cross a factor of roughly two instead decline towards their high-duplication end, and on EU CC the larger contexts sit above the smaller ones at a matched duplication factor without escaping that decline. This is consistent with a simple reading: at a fixed ratio, a larger context multiplies duplicated fraud rows, so high-ratio cells scale the pathology with size while low-ratio cells scale information.

# Ensemble size sweep {#sec-app-nest .appendix}

```{r}
#| label: nest-setup
#| include: false

nest_cells <- c2_grid |>
  filter(fraud_ratio == 0.10, context_size_req == 50000) |>
  summarise(pr_auc = mean(pr_auc), .by = c(dataset, model, n_estimators)) |>
  mutate(model = c(tabiclv2 = "TabICLv2", tabpfn_3 = "TabPFN v3")[model])

nest_v <- function(d, m, n) {
  sprintf("%.3f", nest_cells$pr_auc[nest_cells$dataset == d &
                                      nest_cells$model == m &
                                      nest_cells$n_estimators == n])
}
nest_gap <- nest_cells |>
  summarise(gap = pr_auc[n_estimators == 8] - pr_auc[n_estimators == 4],
            .by = c(dataset, model))
nest_gap_max <- max(nest_gap$gap)
```

The main experiments have capped the number of ensemble passes at n_estimators = 4, a settled design choice for tractability and keeping the GPU-usage lower than having multiple estimators. This sweep verifies that the choice didn't have substantial effects to the performance of the models. It crosses n_estimators over 1, 2, 4, 8, 10 and 12 at the fixed plateau recipe of @sec-app-c2-sweep (10 percent fraud ratio, 50,000-row requested context), on both of the frontier TFMs and all five formal datasets. The original n = 4 cells are shared with the sweep. Within each dataset-seed unit the context rows are identical across n_estimators values, since the sampling does not depend on it, so the sweep isolates the effect of ensembling on a fixed context.

@fig-app-nest shows the outcome. The curves are relatively flat from n = 2 upward and saturate by n = 8. The further increases of n = 10 and n = 12 add virtually nothing performancewise. The settled n = 4 gives up at most about `r sprintf("%.3f", nest_gap_max)` PR-AUC against n = 8 (EU CC with TabPFN v3) and typically at most 0.002. Only a single pass is visibly worse where ensembling matters. On PaySim, TabICLv2 scores `r nest_v("paysim", "TabICLv2", 1)` at n = 1 against `r nest_v("paysim", "TabICLv2", 8)` at saturation of n = 8, and on EU CC, TabPFN v3 scores `r nest_v("eu_cc", "TabPFN v3", 1)` against `r nest_v("eu_cc", "TabPFN v3", 8)` respecticely.

Predict time scales roughly linearly with n_estimators, so the settled n = 4 halves inference cost against the package default of 8 at a negligible accuracy cost. @sec-app-inference quantifies this trade-off with the recorded timing columns.

```{r}
#| label: fig-app-nest
#| fig-cap: "Seed-mean PR-AUC against the number of ensemble passes at the fixed plateau recipe (10 % fraud ratio, 50k context; the vertical scale differs per panel). Curves saturate by n = 8, and the main grid's n_estimators = 4 sits within about 0.006 of the saturated value everywhere."
#| fig-width: 6.2
#| fig-height: 4.2
#| warning: false
#| message: false

nest_cells |>
  ggplot(aes(n_estimators, pr_auc, colour = model)) +
  geom_line(linewidth = 0.6) +
  geom_point(size = 1.2) +
  facet_wrap(~dataset, nrow = 2, scales = "free_y") +
  scale_colour_manual(values = c("TabPFN v3" = "#eb6834",
                                 "TabICLv2" = "#4a3aa7"), name = NULL) +
  scale_x_continuous(breaks = c(1, 2, 4, 8, 10, 12)) +
  labs(x = "Ensemble passes (n_estimators)", y = "PR-AUC") +
  theme_minimal(base_size = 10) +
  theme(legend.position = "bottom", panel.grid.minor = element_blank())
```

# Ratio-matched relevance ablation {#sec-app-relevance-ablation .appendix}

Neither grid contrast isolates relevance cleanly. The step from the random baseline to kNN retrieval varies the selection mechanism but lets the realised fraud ratio drift with the local neighbourhood of each cluster, so a ratio effect and a relevance effect are entangled. The step from plain retrieval to RAP changes ratio enforcement and the retrieval mode at the same time, so it is not a single-factor comparison either. The ablation described here removes both confounds by crossing the two selection mechanisms independently, at a matched ratio, context size and granularity.

The design is a 2x2 over selection mechanisms: legitimate rows drawn either at random or by kNN, crossed with fraud rows drawn either at random or by kNN. This yields four arms: RR (random legitimate rows, random fraud rows), KR (kNN legitimate rows, random fraud rows), RK (random legitimate rows, kNN fraud rows) and KK (kNN both). All four arms run per group at a fixed granularity of G = 32 k-means clusters of the scored test set. Granularity is fixed rather than silhouette-selected here, so grouping is held constant across arms and cannot itself explain a difference between them. Retrieval uses the cosine metric, the requested context size is 50,000 rows, and the fraud ratio is swept over 5 percent and 10 percent, filled with the same with-replacement top-up policy as the stratified and RAP conditions. The ablation runs on TabICLv2 and TabPFN v3, on EU CC, BankSim and PaySim, at three seeds. RR reproduces the stratified condition's random selection, operated at a set ratio rather than the 50/50 target and with the same with-replacement policy as the sweep, under per-group refitting, and KK is the class-conditional retrieval core of RAP; each single-letter flip from one of these arms isolates one selector. Because RR runs at the same G = 32 as the other three arms, comparing it against the global-context sweep of @sec-app-c2-sweep additionally isolates the effect of per-group granularity itself, separately from the effect of any selector.

The ablation is additionally run at a natural ratio level, defined per dataset as the training-split fraud prevalence, for the RR and KK arms only. This supplies the clean form of the random-versus-retrieval contrast: both arms see the same natural class ratio and the same grouping, so only the selection mechanism differs between them. At the natural level the fraud quota sits below the unique fraud count by construction, so no duplication occurs. <!-- TODO: inline R for the realised natural ratios per dataset -->

Each contrast in this design answers a distinct question. The ratio effect is addressed in @sec-methods-conditions. Granularity is addressed by RR against the global sweep of @sec-app-c2-sweep. Legitimate-selector relevance is addressed by RK against KK and by RR against KR. Fraud-selector relevance is addressed by RR against RK and by KR against KK. The selection mechanism at natural prevalence, free of ratio enforcement, is addressed by RR against KK at the natural level.

# Inference-time scaling {#sec-app-inference .appendix}

```{r}
#| label: inftime-setup
#| include: false

# Sweep-side objects (reuse c2_grid from c2-app-setup). The three sweep
# datasets carry 7 (banksim, paysim) or 30 (eu_cc) features; the
# n_estimators sweep adds baf (30) and fifar (29) rows at the 50k recipe.
inft_disp <- c(tabpfn_v2 = "TabPFN v2", tabpfn_25 = "TabPFN 2.5",
               tabpfn_26 = "TabPFN 2.6", tabpfn_3 = "TabPFN v3",
               tabiclv2 = "TabICLv2")
inft_col <- c("TabPFN v3" = "#eb6834", "TabICLv2" = "#4a3aa7")

inft_ctx <- c2_grid |>
  filter(n_estimators == 4, dataset %in% c("banksim", "eu_cc", "paysim")) |>
  mutate(s_per_1k = predict_s / (n_test_scored / 1000)) |>
  summarise(predict_s = mean(predict_s), s_per_1k = mean(s_per_1k),
            .by = c(model, context_size_req)) |>
  mutate(model = inft_disp[model])

inft_at <- function(m, s, col = "predict_s") {
  inft_ctx[[col]][inft_ctx$model == m & inft_ctx$context_size_req == s]
}
inft_ratio_v3 <- inft_at("TabPFN v3", 2e5) / inft_at("TabPFN v3", 5e4)
inft_ratio_icl <- inft_at("TabICLv2", 2e5) / inft_at("TabICLv2", 5e4)
inft_fit_max <- c2_grid |>
  filter(dataset %in% c("banksim", "eu_cc", "paysim")) |>
  summarise(m = max(fit_s)) |>
  pull(m)

inft_nest <- c2_grid |>
  filter(fraud_ratio == 0.10, context_size_req == 50000) |>
  summarise(predict_s = mean(predict_s), .by = c(model, n_estimators)) |>
  mutate(model = inft_disp[model])
inft_pass <- c2_grid |>
  filter(fraud_ratio == 0.10, context_size_req == 50000) |>
  summarise(per_pass = mean(predict_s / n_estimators), .by = model)
inft_pass_icl <- inft_pass$per_pass[inft_pass$model == "tabiclv2"]
inft_pass_v3 <- inft_pass$per_pass[inft_pass$model == "tabpfn_3"]

inft_feat_incr <- c2_grid |>
  filter(n_estimators == 4, context_size_req == 50000) |>
  mutate(s_per_1k = predict_s / (n_test_scored / 1000),
         width = ifelse(dataset %in% c("banksim", "paysim"),
                        "narrow", "wide")) |>
  summarise(s_per_1k = mean(s_per_1k), .by = c(model, dataset, width)) |>
  summarise(s_per_1k = mean(s_per_1k), .by = c(model, width)) |>
  tidyr::pivot_wider(names_from = width, values_from = s_per_1k) |>
  mutate(incr = wide / narrow - 1)
inft_incr_icl <- inft_feat_incr$incr[inft_feat_incr$model == "tabiclv2"]
inft_incr_v3 <- inft_feat_incr$incr[inft_feat_incr$model == "tabpfn_3"]

# Formal-grid rows (results/runs). bind_rows fails on the mixed-type
# `metric` column, so select the needed columns per file (as in the
# results chapter).
inft_cols <- c("dataset", "model", "condition", "seed", "n_estimators",
               "context_size", "n_test_scored", "fit_s", "predict_s")
inft_runs <- list.files("results/runs", pattern = "\\.parquet$",
                        full.names = TRUE) |>
  lapply(\(f) nanoparquet::read_parquet(f)[, inft_cols]) |>
  bind_rows()

inft_ftms <- names(inft_disp)
inft_models <- inft_runs |>
  filter(dataset != "ai_banking", model %in% inft_ftms,
         condition %in% c("C1", "C2")) |>
  mutate(s_per_1k = predict_s / (n_test_scored / 1000)) |>
  summarise(context = mean(context_size), s_per_1k = mean(s_per_1k),
            fit_s = mean(fit_s), .by = model) |>
  mutate(chunk = c(tabpfn_v2 = 2048, tabpfn_25 = 1024, tabpfn_26 = 1024,
                   tabpfn_3 = 1024, tabiclv2 = 2048)[model],
         model = inft_disp[model]) |>
  arrange(s_per_1k)
inft_r_25v3 <- inft_models$s_per_1k[inft_models$model == "TabPFN 2.5"] /
  inft_models$s_per_1k[inft_models$model == "TabPFN v3"]
inft_r_v3icl <- inft_models$s_per_1k[inft_models$model == "TabPFN v3"] /
  inft_models$s_per_1k[inft_models$model == "TabICLv2"]
inft_r_iclv2 <- inft_models$s_per_1k[inft_models$model == "TabICLv2"] /
  inft_models$s_per_1k[inft_models$model == "TabPFN v2"]
inft_r_v3v2 <- inft_models$s_per_1k[inft_models$model == "TabPFN v3"] /
  inft_models$s_per_1k[inft_models$model == "TabPFN v2"]
inft_r_sweep <- inft_at("TabPFN v3", 5e4, "s_per_1k") /
  inft_at("TabICLv2", 5e4, "s_per_1k")

inft_total <- inft_runs |>
  filter(model %in% inft_ftms) |>
  summarise(n = n(), hours = sum(fit_s + predict_s) / 3600)
inft_xgb_fit <- inft_runs |>
  filter(model == "xgboost", dataset != "ai_banking") |>
  summarise(m = mean(fit_s)) |>
  pull(m)

# same-cell cross-directory comparison (tabpfn_3, banksim, 50k, n = 4)
inft_xdir_grid <- inft_runs |>
  filter(model == "tabpfn_3", dataset == "banksim", condition == "C2") |>
  summarise(m = mean(predict_s)) |>
  pull(m)
inft_xdir_sweep <- c2_grid |>
  filter(model == "tabpfn_3", dataset == "banksim", fraud_ratio == 0.10,
         n_estimators == 4, context_size_req == 50000) |>
  summarise(m = mean(predict_s)) |>
  pull(m)
inft_xdir_ratio <- inft_xdir_grid / inft_xdir_sweep

cap_inftime_fig <- glue(
  "Predict-time scaling in the sweep of @sec-app-c2-sweep (16,384-row ",
  "predict chunks; absolute levels are not comparable to the formal grid). ",
  "Left: mean predict seconds per 1,000 scored rows against requested ",
  "context size at n_estimators = 4, averaged over the three sweep ",
  "datasets, all ratios and seeds (log-log axes; the dashed line marks ",
  "linear scaling). Right: mean predict seconds against ensemble passes at ",
  "the 10 % ratio and 50,000-row context, averaged over all five formal ",
  "datasets and seeds; per-pass cost is roughly constant at about ",
  "{sprintf('%.1f', inft_pass_icl)} s (TabICLv2) and ",
  "{sprintf('%.1f', inft_pass_v3)} s (TabPFN v3)."
)
```

For a TFM essentially all compute happens at the prediction time. The gradient-boosted baselines invert this profile. They pay once at training XGBoost mean fit `r sprintf("%.1f", inft_xgb_fit)` s on the formal grid and then predict in milliseconds. (CatBoost was pinned to a single thread for stability in this pipeline, so its fit times are not representative and are omitted.) Summed over all `r inft_total$n` TFM runs in this thesis inference cost about `r sprintf("%.0f", inft_total$hours)` GPU-hours on the single consumer GPU throughout (NVIDIA RTX 5070, 12 GB VRAM).

Two result sets supply the timings: the sweep of @sec-app-c2-sweep for the within-model scaling factors, and the formal experiment grid for the five-model comparison, restricted to the random and stratified conditions. Retrieval-condition rows record per-group sums under a different refitting regime and are excluded. Absolute times are not comparable between the two sets. The grid runner predicts in small GPU-memory chunks of 1,024 test rows for TabPFN 2.5, 2.6 and v3 and 2,048 for TabPFN v2 and TabICLv2, and every chunk re-encodes the fixed context. The sweep driver uses 16,384-row chunks, which amortise the context encoding over eight to sixteen times more rows, depending on the model. A near-identical cell (TabPFN v3, BankSim, 50,000-row context, n_estimators = 4; the grid's stratified condition against the sweep's 10 % ratio cell) costs `r sprintf("%.0f", inft_xdir_grid)` s in the grid against `r sprintf("%.0f", inft_xdir_sweep)` s in the sweep, roughly `r sprintf("%.0f", inft_xdir_ratio)`-fold more. Within-set comparisons remain valid; the grid times are conservative upper bounds on deployment cost, while the sweep times are closer to a throughput-oriented setting.

Context size is the dominant driver, and its effect is superlinear as shown in the left panel of @fig-app-inftime. At 4 estimatros the mean predict time of TabPFN v3 rises from `r sprintf("%.1f", inft_at("TabPFN v3", 1e4))` s at a 10,000-row context to `r sprintf("%.0f", inft_at("TabPFN v3", 5e4))` s at 50,000 rows and `r sprintf("%.0f", inft_at("TabPFN v3", 2e5))` s at 200,000 rows. TabICLv2 rises from `r sprintf("%.1f", inft_at("TabICLv2", 1e4))` s to `r sprintf("%.0f", inft_at("TabICLv2", 5e4))` s and `r sprintf("%.0f", inft_at("TabICLv2", 2e5))` s over the same range. Quadrupling the context from 50,000 to 200,000 rows therefore costs about `r sprintf("%.0f", inft_ratio_v3)` times the predict time on TabPFN v3 and about `r sprintf("%.0f", inft_ratio_icl)` times on TabICLv2, well above the dashed linear reference in the figure.

Ensemble size scales cleanly, shown in the right panel if @fig-app-inftime. At the 10 % ratio and 50,000-row context, predict time is linear in n_estimators with a roughly constant per-pass cost of about `r sprintf("%.1f", inft_pass_icl)` s for TabICLv2 and `r sprintf("%.1f", inft_pass_v3)` s for TabPFN v3. The accuracy side of this trade-off is in @sec-app-nest: the settled n = 4 halves predict time against the package default of 8 at a PR-AUC cost of at most about 0.006.

Feature count matters far less. Per-row cost is normalised per 1,000 scored rows, since EU CC scores about 57,000 test rows against about 31,000 to 32,000 elsewhere. Moving from the 7-feature datasets to the 29- and 30-feature ones at the 50,000-row context then raises per-row cost by about `r pct(inft_incr_v3, 1)` for TabPFN v3 and `r pct(inft_incr_icl, 1)` for TabICLv2, with EU CC contributing most of the increase. The effect is modest next to context size.

@tbl-app-inftime-models compares the five TFMs. TabPFN v2 is the cheapest per scored row, about `r sprintf("%.1f", inft_r_iclv2)` times below TabICLv2 and roughly `r sprintf("%.0f", inft_r_v3v2)` times below TabPFN v3, mostly because its checkpoint caps the context at 10,000 rows. At near-identical budgets TabPFN v3 costs about `r sprintf("%.1f", inft_r_v3icl)` times more per scored row than TabICLv2, though the differing predict chunks (1,024 against 2,048 rows) inflate this ratio. In the sweep, where both models share one chunk size, the same comparison at the 50,000-row context gives about `r sprintf("%.1f", inft_r_sweep)` times. TabPFN 2.5 and 2.6 run at the same budget and chunk size as v3 and still cost about `r sprintf("%.1f", inft_r_25v3)` times more, which suggests a genuinely heavier architecture rather than a chunking artifact. Given the caveats above, these ratios should be read as indicative.

```{r}
#| label: fig-app-inftime
#| fig-cap: !expr cap_inftime_fig
#| fig-width: 6.2
#| fig-height: 3.4
#| warning: false
#| message: false

inft_ref_a <- inft_at("TabICLv2", 1e4, "s_per_1k") / 1e4
inft_ref <- data.frame(context_size_req = c(1e4, 2e5)) |>
  mutate(s_per_1k = inft_ref_a * context_size_req)

p_inft_ctx <- inft_ctx |>
  ggplot(aes(context_size_req, s_per_1k, colour = model)) +
  geom_line(data = inft_ref, aes(colour = NULL), linetype = "dashed",
            colour = "grey55", linewidth = 0.4, show.legend = FALSE) +
  annotate("text", x = 1.05e5, y = inft_ref_a * 1.05e5 * 0.55,
           label = "linear (slope 1)", colour = "grey40", size = 2.6,
           angle = 26) +
  geom_line(linewidth = 0.6) +
  geom_point(size = 1.2) +
  scale_colour_manual(values = inft_col, name = NULL) +
  scale_x_log10(breaks = c(1e4, 2e4, 5e4, 1e5, 2e5),
                labels = label_number(scale = 1e-3, suffix = "k")) +
  scale_y_log10() +
  labs(x = "Requested context size (rows)",
       y = "Predict s per 1,000 scored rows") +
  theme_minimal(base_size = 10) +
  theme(legend.position = "bottom", panel.grid.minor = element_blank())

p_inft_nest <- inft_nest |>
  ggplot(aes(n_estimators, predict_s, colour = model)) +
  geom_line(linewidth = 0.6) +
  geom_point(size = 1.2) +
  scale_colour_manual(values = inft_col, name = NULL) +
  scale_x_continuous(breaks = c(1, 2, 4, 8, 10, 12)) +
  labs(x = "Ensemble passes (n_estimators)", y = "Predict s") +
  theme_minimal(base_size = 10) +
  theme(legend.position = "bottom", panel.grid.minor = element_blank())

(p_inft_ctx | p_inft_nest) + plot_layout(guides = "collect") &
  theme(legend.position = "bottom")
```

```{r}
#| label: tbl-app-inftime-models
#| tbl-cap: "Mean inference time per TFM over the formal-grid random- and stratified-condition runs. Absolute levels reflect the grid runner’s small predict chunks, which re-encode the context on every chunk."

inft_models |>
  select(model, context, chunk, s_per_1k, fit_s) |>
  gt() |>
  cols_label(
    model = "Model",
    context = "Context budget",
    chunk = "Predict chunk",
    s_per_1k = "Predict s / 1,000 rows",
    fit_s = "Mean fit s"
  ) |>
  fmt_number(columns = c(context, chunk), decimals = 0, use_seps = TRUE) |>
  fmt_number(columns = c(s_per_1k, fit_s), decimals = 2)
```
